---
title: "Computing SDG 5.3.1 Using MICS Data in R"
author: "Sinovia Moonie, Statistician/Lecturer, Statistical Institute for Asia and the Pacific"
subtitle: '*A simple report*'
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    code_download: true
  word_document:
    toc: no
  pdf_document:
    toc: yes
always_allow_html: true  # <-- bypass errors caused by interactive HTML
---


```{r setup, include=FALSE}
# We define the options for the chunks globaly here

knitr::opts_chunk$set( message = FALSE, warning = FALSE, results =TRUE, echo = TRUE) 

```

# Introduction

This tutorial demonstrates how to compute **SDG indicator 5.3.1: Child Marriage** using microdata from the **Multiple Indicator Cluster Survey (MICS)** for Lao PDR (2023).

SDG 5.3.1 measures:

> *The proportion of women aged 20–24 years who were first married or in union before age 18.*

The tutorial introduces a simple workflow for:
- importing survey data,
- exploring variables,
- filtering the relevant population,
- calculating weighted indicators,
- producing disaggregated estimates,
- visualizing the data,
- and exporting results.

# Step 1: Install the appropriate R packages 

*Install packages only once.*

We first load the packages needed for this exercise.

The package `haven` is used to import SPSS files (`.sav`), while `tidyverse` provides tools for data manipulation, visualization, and analysis.

*Hint:* we use here a package manager `pacman` (another package). Once installed and loaded, it can facilitate the use of a list of packages  

```{r packages}
# We use pacman here 
# install.packages("pacman")  # <-- Run only once

library(pacman)
pacman::p_load(
  haven,
  tidyverse
)
```

# 2. Define File Paths

It is good practice to define folders where your outputs will be stored. 

```{r}

# input = file.path(base_path, "Data")
# outputFolder = file.path(base_path, "Output")
```

# 3. Import the Data

The questionnaire dataset (`wm.sav`) is imported into R using `read_spss()`.

```{r import}
#Create a data frame called df which stores the MICS data
rawdf <- read_spss("../Data/wm.sav")
```

# 4. Explore the Dataset

This dataset contains information collected from women interviewed during the survey. it has `r nrow(df)` observations and `r ncol(df)` variables 


Although the dataset contains many variables, we only need a few for this computation:

| Variable | Description |
|---|---|
| WM17 | Interview completion status |
| WB4 | Age of woman |
| WAGEM | Age at first marriage |
| HH6 |  Area of residence    |
| wmweight | Sampling weight |


##  Suggestion for Sinovia
 > Let us select the variable of interest in a new, smaller data set
 
```{r}
df <- rawdf %>%
  select(WM17, WB4, WAGEM,HH6, wmweight)

```
 

# 5. Explore Variable Structure

## Check Variable Classes for WB4 and WAGEM

```{r type}
df %>% 
  select(WB4, WAGEM) %>% 
  map(class)
```

## Display an overview of the data

```{r head}
head(df)
```

## List Variable Names

We have here `r ncol(df)` variables: *`r names(df)`*

# 6. Explore Interview Status

The variable `WM17` indicates the completion status of the interview. We may want to check the number of successful interviews

```{r InfoInterview }
str(df$WM17)
```


```{r taleInterview}
table(df$WM17)
```

## Bar Chart of Interview Status

```{r BarInterview}
df %>%
  count(WM17) %>%  # We use the count function to have the values
  mutate(
    pct = n / sum(n) * 100,
    label = paste0(round(pct, 1), "%")
  ) %>%
  ggplot(aes(x = "", y = n, fill = factor(WM17))) +
  geom_col(width = 0.3) +
  geom_text(
    data = \(x) filter(x, WM17 == 1),
    aes(label = label),
    position = position_stack(vjust = 0.5),
    color = "white"
  ) +
  coord_flip() +
  labs(
    title = "Interview Status",
    x = "",
    y = "Number of Women",
    fill = "WM17"
  ) +
  
  theme_minimal()
```

# 7. Explore Age Variables

## Summary Statistics

```{r summaryAge}


df %>% 
  select(WB4) %>% 
  summary()

# summary(df$WB4)

```

## Histogram of Women's Age

```{r HistAge}
df %>%
ggplot() +
  aes(x = WB4) +
  geom_histogram(binwidth = 1, fill = "lightblue") +
  labs(
    title = "Distribution of Women's Age",
    x = "Age",
    y = "Frequency"
  ) +
  theme_minimal()
```

# 8. Explore Sampling Weights

Sampling weights are used to produce representative estimates at the national level.

```{r}
summary(df$wmweight)
```

## Distribution of Sampling Weights

```{r}
ggplot(df, aes(x = wmweight)) +
  geom_histogram(bins = 30 ,
                 fill = "lightgrey", 
                 colour = "white") +
  labs(
    title = "Distribution of Sampling Weights",
    x = "Sampling Weight",
    y = "Frequency"
  ) +
  theme_minimal()
```

# 9. Explore Age at First Marriage

```{r}
summary(df$WAGEM)
```

## Histogram of Age at First Marriage

```{r}
ggplot(df, aes(x = WAGEM)) +
  geom_histogram(binwidth = 1,
                 fill = "pink", 
                 colour = "white") +
  geom_vline(xintercept = 16,
             colour = "darkred",
             linewidth =1) +
  annotate("text", x = 9, y = 1800,
           label = "Child marriage if <16",
           color = "darkred",
           size = 4)+
  labs(
    title = "Age at First Marriage",
    x = "Age",
    y = "Frequency"
  ) +
  theme_minimal()
```

# 10. Select the Population of Interest

SDG 5.3.1 is calculated for women aged **20–24 years** who completed the interview.

```{r}
wm_20_24 <- df %>%
  filter(
    WM17 == 1,
    WB4 >= 20,
    WB4 < 25
  )
```

We can compute the number of women included in the analysis.

```{r}
nrow(wm_20_24)
```

# 11. Compute SDG 5.3.1

The indicator is calculated as:

$$
\text{SDG 5.3.1} =
\frac{\text{Women married before age 18}}
{\text{Total women aged 20–24}}
\times 100
$$

Because MICS is a sample survey, we use the sampling weights (`wmweight`) in both the numerator and denominator.

```{r}
sdg_5_3_1 <- wm_20_24 %>%
  summarise(
    indicator = "SDG 5.3.1",
    denominator = sum(wmweight, na.rm = TRUE),
    numerator   = sum(wmweight[WAGEM < 18], na.rm = TRUE),
    SDG_5_3_1    = 100 * numerator / denominator
  )

sdg_5_3_1
```

# 12. Disaggregation by Location

Indicators are often disaggregated to study inequalities between groups.

Here we calculate SDG 5.3.1 separately for urban and rural areas.

```{r}
sdg_5_3_1_location <- wm_20_24 %>%
  group_by(HH6) %>%
  summarise(
    denominator = sum(wmweight, na.rm = TRUE),
    numerator   = sum(wmweight[WAGEM < 18], na.rm = TRUE),
    proportion  = numerator / denominator
  )

sdg_5_3_1_location
```

## Convert Proportions to Percentages

```{r}
sdg_5_3_1_location <- sdg_5_3_1_location %>%
  mutate(proportion = proportion * 100)

sdg_5_3_1_location
```

# 13. Visualize Urban and Rural Differences

```{r}
ggplot(
  sdg_5_3_1_location,
  aes(x = factor(HH6), y = proportion)
) +
  geom_col() +
  labs(
    title = "SDG 5.3.1 by Location",
    x = "Location (1 = Urban, 2 = Rural)",
    y = "Percent"
  ) +
  theme_minimal()
```

# 14. Export the Results

The final results are saved as a CSV file.

```{r}
results <- sdg_5_3_1 %>%
  mutate(
    country = "Lao PDR",
    year = 2023,
    unit = "Percent"
  )

write_csv(results, file.path(output, "SDG_5_3_1_results.csv"))
```

# Conclusion

This tutorial illustrated a complete workflow for computing an SDG indicator using household survey microdata in R.

The workflow included:
- importing SPSS data,
- exploring variables,
- producing descriptive statistics,
- visualizing distributions,
- filtering the target population,
- calculating weighted indicators,
- producing disaggregated estimates,
- and exporting the results.

The same general approach can be adapted to many other SDG indicators derived from survey data.