---
title: "Computing SDG 5.3.1 Using MICS Data in R"
author: "Training Material"
output: html_document
---

# Introduction

This tutorial demonstrates how to compute **SDG indicator 5.3.1: Child Marriage** using microdata from the **Multiple Indicator Cluster Survey (MICS)** for Lao PDR (2023).

SDG 5.3.1 measures:

> *The proportion of women aged 20–24 years who were first married or in union before age 18.*

The tutorial introduces a simple workflow for:
- importing survey data,
- exploring variables,
- filtering the relevant population,
- calculating weighted indicators,
- producing disaggregated estimates,
- visualizing the data,
- and exporting results.

# 1. Load Required Packages

We first load the packages needed for this exercise.

The package `haven` is used to import SPSS files (`.sav`), while `tidyverse` provides tools for data manipulation, visualization, and analysis.

```{r}
# install.packages("pacman")

pacman::p_load(
  haven,
  tidyverse
)
```

# 2. Define File Paths

To make scripts easier to maintain and reuse, it is good practice to define folders only once.

```{r}
base_path = "C:/Users/un157/OneDrive - United Nations/Desktop/Practical/LaoPDR"

input = file.path(base_path, "Data")
output = file.path(base_path, "Output")
```

# 3. Import the Data

The women's questionnaire dataset (`wm.sav`) is imported into R using `read_spss()`.

```{r}
df <- read_spss(file.path(input, "wm.sav"), user_na = TRUE)
```

# 4. Explore the Dataset

This dataset contains information collected from women interviewed during the survey.

The following chunk computes the number of observations (rows) and variables (columns) in the dataset.

```{r}
nrow(df)

ncol(df)
```

We can also display both values together.

```{r}
cat(
  "The dataset contains",
  nrow(df),
  "observations and",
  ncol(df),
  "variables."
)
```

Although the dataset contains many variables, we only need a few for this computation:

| Variable | Description |
|---|---|
| WM17 | Interview completion status |
| WB4 | Age of woman |
| WAGEM | Age at first marriage |
| wmweight | Sampling weight |

# 5. Explore Variable Structure

## Check Variable Classes

```{r}
class(df$WB4)

class(df$WAGEM)
```

## Display Dataset Structure

```{r}
glimpse(df)
```

## List Variable Names

```{r}
names(df)
```

# 6. Explore Interview Status

The variable `WM17` indicates the completion status of the interview.

```{r}
str(df$WM17)

table(df$WM17)
```

## Bar Chart of Interview Status

```{r}
df %>%
  count(WM17) %>%
  ggplot(aes(x = factor(WM17), y = n)) +
  geom_col() +
  labs(
    title = "Interview Status",
    x = "WM17",
    y = "Number of Women"
  ) +
  theme_minimal()
```

# 7. Explore Age Variables

## Summary Statistics

```{r}
summary(df$WB4)

range(df$WB4, na.rm = TRUE)
```

## Histogram of Women's Age

```{r}
ggplot(df, aes(x = WB4)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Distribution of Women's Age",
    x = "Age",
    y = "Frequency"
  ) +
  theme_minimal()
```

# 8. Explore Sampling Weights

Sampling weights are used to produce representative estimates at the national level.

```{r}
summary(df$wmweight)
```

## Distribution of Sampling Weights

```{r}
ggplot(df, aes(x = wmweight)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Sampling Weights",
    x = "Sampling Weight",
    y = "Frequency"
  ) +
  theme_minimal()
```

# 9. Explore Age at First Marriage

```{r}
summary(df$WAGEM)
```

## Histogram of Age at First Marriage

```{r}
ggplot(df, aes(x = WAGEM)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Age at First Marriage",
    x = "Age",
    y = "Frequency"
  ) +
  theme_minimal()
```

# 10. Select the Population of Interest

SDG 5.3.1 is calculated for women aged **20–24 years** who completed the interview.

```{r}
wm_20_24 <- df %>%
  filter(
    WM17 == 1,
    WB4 >= 20,
    WB4 < 25
  )
```

We can compute the number of women included in the analysis.

```{r}
nrow(wm_20_24)
```

# 11. Compute SDG 5.3.1

The indicator is calculated as:

$$
\text{SDG 5.3.1} =
\frac{\text{Women married before age 18}}
{\text{Total women aged 20–24}}
\times 100
$$

Because MICS is a sample survey, we use the sampling weights (`wmweight`) in both the numerator and denominator.

```{r}
sdg_5_3_1 <- wm_20_24 %>%
  summarise(
    indicator = "SDG 5.3.1",
    denominator = sum(wmweight, na.rm = TRUE),
    numerator   = sum(wmweight[WAGEM < 18], na.rm = TRUE),
    SDG_5_3_1    = 100 * numerator / denominator
  )

sdg_5_3_1
```

# 12. Disaggregation by Location

Indicators are often disaggregated to study inequalities between groups.

Here we calculate SDG 5.3.1 separately for urban and rural areas.

```{r}
sdg_5_3_1_location <- wm_20_24 %>%
  group_by(HH6) %>%
  summarise(
    denominator = sum(wmweight, na.rm = TRUE),
    numerator   = sum(wmweight[WAGEM < 18], na.rm = TRUE),
    proportion  = numerator / denominator
  )

sdg_5_3_1_location
```

## Convert Proportions to Percentages

```{r}
sdg_5_3_1_location <- sdg_5_3_1_location %>%
  mutate(proportion = proportion * 100)

sdg_5_3_1_location
```

# 13. Visualize Urban and Rural Differences

```{r}
ggplot(
  sdg_5_3_1_location,
  aes(x = factor(HH6), y = proportion)
) +
  geom_col() +
  labs(
    title = "SDG 5.3.1 by Location",
    x = "Location (1 = Urban, 2 = Rural)",
    y = "Percent"
  ) +
  theme_minimal()
```

# 14. Export the Results

The final results are saved as a CSV file.

```{r}
results <- sdg_5_3_1 %>%
  mutate(
    country = "Lao PDR",
    year = 2023,
    unit = "Percent"
  )

write_csv(results, file.path(output, "SDG_5_3_1_results.csv"))
```

# Conclusion

This tutorial illustrated a complete workflow for computing an SDG indicator using household survey microdata in R.

The workflow included:
- importing SPSS data,
- exploring variables,
- producing descriptive statistics,
- visualizing distributions,
- filtering the target population,
- calculating weighted indicators,
- producing disaggregated estimates,
- and exporting the results.

The same general approach can be adapted to many other SDG indicators derived from survey data.