---
title: "Computing SDG 5.3.1 Using MICS Data in R"
subtitle: '*A simple report*'
author: "Your Name"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    code_download: true
  word_document:
    toc: no
always_allow_html: true  # <-- bypass errors caused by interactive HTML (if any)
---


```{r setup, include=FALSE}
# We define the options for the chunks globaly here

knitr::opts_chunk$set( message = FALSE,
                       warning = FALSE,
                       results =TRUE
                       
                       )

```

```{r Country, eval=TRUE, include=FALSE}
# Since we may want to use other data set from other countries/years
# Let us define the Country and year as a parameter
MyCountry <- "Lao PDR"
DataYear <-  "2023"

```

# Introduction

This tutorial demonstrates how to compute **SDG indicator 5.3.1: Child Marriage** using microdata from the **Multiple Indicator Cluster Survey (MICS)** for `r MyCountry`  (`r DataYear`).

SDG 5.3.1 measures:

> *The proportion of women aged 20–24 years who were first married or in union before age 18.*

The tutorial introduces a simple workflow for:

- importing survey data,
- exploring variables,
- filtering the relevant population,
- calculating weighted indicators,
- producing disaggregated estimates,
- visualizing the data,
- and exporting results.

# 1: Install the appropriate R packages 

*Install packages only once.*

We first load the packages needed for this exercise.

The package `haven` is used to import SPSS files (`.sav`), while `tidyverse` provides tools for data manipulation, visualization, and analysis.

*Hint:* we use here a package manager `pacman` (another package). Once installed and loaded, it can facilitate the use of a list of packages  

```{r packages}
# We use pacman here 
#install.packages("pacman")  # <-- Run only once

library(pacman)
pacman::p_load(
  haven,
  kableExtra, 
  modelsummary,
  tidyverse
)
```


# 2. Import the Data

The full questionnaire dataset (`wm.sav`) is imported into R using `read_spss()`.

```{r import, eval = TRUE}
#Create a data frame called df which stores the MICS data
# This data set will change for a different Country
rawdf <- haven::read_spss("../Data/wm.sav")
```

The raw data set has ***`r nrow(rawdf)`*** observations and  **476** variables. 

# 3. Explore the Dataset

This dataset contains information collected from women interviewed during the survey. it has `r nrow(rawdf)` observations and `r ncol(rawdf)` variables 


Although the dataset contains many variables, we only need a few for this computation:

| Variable | Description |
|---|---|
| WM17 | Interview completion status |
| WB4 | Age of woman |
| WAGEM | Age at first marriage |
| HH6 |  Area of residence    |
| wmweight | Sampling weight |


##  Variable selection
 > Let us select the variable of interest in a new, smaller data set with only the variables of interest, that is "WM17", "WB4", "WAGEM", "HH6",and "wmweight".
 
 
```{r subset}
df <- rawdf %>%
  select("WM17", "WB4", "WAGEM", "HH6","wmweight")

```

We have now a new data set  `df` with still **22512 ** observations   but only  **5** variables! Let us use some fancy package to get an overview of the values of these 5 variables: 


 
```{r Summary }
# We can use the package modelsummary for a nice descriptive table  
datasummary_skim(df , type = "numeric" )
```


# 5. Explore Variable Structure


> Let us change the type of the variables to match their original nature. 

The variable `HH6` is considered as numerical in the original data set, while it encodes the Area of residence (Rural, Urban). This should be a ***factor***. We should modify its type.  

```{r typeHH6}
# The variable HH6 should be a factor
# We use https://microdata.worldbank.org/index.php/catalog/4168/data-dictionary/F4?file_name=hh.sav 

df <- df %>%
  mutate(HH6 = factor(HH6,
                       levels = c(1,2,3),
                       labels = c("Urban", "Rural", "Unknown")
                       )
         )
```



## Check Variable Classes for WB4 and WAGEM

```{r type}
df %>% 
  map(class)
```

## Display an overview of the data

```{r head}
head(df)
```



# 6. Explore Interview Status

The variable `WM17` indicates the completion status of the interview. We may want to check the number of successful interviews

```{r InfoInterview }
str(df$WM17)
```


```{r tableInterview}
table(df$WM17)
```

## Bar Chart of Interview Status

```{r BarInterview}
df %>%
  count(WM17) %>%  # We use the count function to have the values
  mutate(
    pct = n / sum(n) * 100,
    label = paste0(round(pct, 1), "%")
  ) %>%
  ggplot(aes(x = "", y = n, fill = factor(WM17))) +
  geom_col(width = 0.3) +
  geom_text(
    data = \(x) filter(x, WM17 == 1),
    aes(label = label),
    position = position_stack(vjust = 0.5),
    color = "white"
  ) +
  coord_flip() +
  labs(
    title = "Interview Status",
    x = "",
    y = "Number of Women",
    caption = paste0("Source: MICS - ", MyCountry,", (", DataYear,")"),
    fill = "Interview result"
  ) +
  
  theme_minimal()
```

# 7. Explore Age Variable

## Summary Statistics


```{r summaryAge}
# Age is the variable WB4

df %>% 
  select(WB4) %>% 
  summary()

# summary(df$WB4)

```

## Histogram of Women's Age

```{r HistAge}
# We can display the distribution with an Histogram

df %>%
ggplot() +
  aes(x = WB4) +
  geom_histogram(binwidth = 1, fill = "lightblue") +
   geom_vline(aes(xintercept = median(WB4, na.rm =TRUE)), 
             color = "red", 
             linetype = "dashed", 
             linewidth = 1) +
  
  labs(
    title = "Distribution of Women's Age",
    x = "Age",
    y = "Frequency", 
    caption = paste0("Source: MICS - ", MyCountry,", (", DataYear,")"),
  ) +
  theme_minimal()
```

# 8. Explore Sampling Weights

Sampling weights are used to produce representative estimates at the national level.

```{r}

df %>%
  select(wmweight) %>%
  summary()
```

## Distribution of Sampling Weights

```{r}
ggplot(df, aes(x = wmweight)) +
  geom_histogram(bins = 30 ,
                 fill = "lightgrey", 
                 colour = "white") +
  labs(
    title = "Distribution of Sampling Weights",
    x = "Sampling Weight",
    y = "Frequency", 
   caption = paste0("Source: MICS - ", MyCountry,", (", DataYear,")")  ) +
  theme_minimal()
```

# 9. Explore Age at First Marriage

```{r}
df %>%
  select(WAGEM) %>%
  summary()
```

## Histogram of Age at First Marriage

```{r}
ggplot(df, aes(x = WAGEM)) +
  geom_histogram(binwidth = 1,
                 fill = "pink", 
                 colour = "white") +
  geom_vline(xintercept = 16,
             colour = "darkred",
             linewidth =1) +
  annotate("text", x = 9, y = 1800,
           label = "Child marriage if <16",
           color = "darkred",
           size = 4)+
  labs(
    title = "Age at First Marriage",
    x = "Age at first marriage",
    y = "Frequency",
    caption = paste0("Source: MICS - ", MyCountry,", (", DataYear,")",
     " - Nb of respondants =", nrow(df) ) 
    ) +
  theme_minimal()
```

# 10. Select the Population of Interest

SDG 5.3.1 is calculated for women aged **20–24 years** who completed the interview. We create a new data set `wm_20_24` with only the population in the definition.  



```{r subset20_24}
wm_20_24 <- df %>%
 filter(
   WB4 >= 20 &   # age greater or equal 20
    WB4 <= 24 &  #  age  lower or equal 24 
     WM17 ==1    # Completed interview
     
 )
  
  
  
```

We can compute the number of women included in the analysis, now only **XXXX** women.



# 11. Compute SDG 5.3.1

The indicator is calculated as:

$$
\text{SDG 5.3.1} =
\frac{\text{Nb of women married before age 18}}
{\text{Total nb women aged 20–24}}
\times 100
$$

Because MICS is a sample survey, we use the sampling weights (`wmweight`) in both the numerator and denominator, so the exact definition becomes.

$$
\text{SDG 5.3.1} =
\frac{\text{ "weighted nb" of women married before age 18}}
{\text{Total "weighted nb" women aged 20–24}}
\times 100
$$
The result can be summarized in this table, where we compute the numerator, the denominator and the ratio (multiplied by 1000):

```{r SDG_5_3_1}
sdg_5_3_1 <- wm_20_24 %>%
  summarise(
    
    denominator = sum(wmweight , na.rm = TRUE),
    numerator = sum(wmweight[WAGEM < 18], na.rm = TRUE),
    indicator = 100 * numerator/denominator
    
  ) 
 
# Making a nice table
kable(sdg_5_3_1,
      digits = 1,
      caption = paste0("SDG 5.3.1 - ",MyCountry,", (", DataYear,")"))
```

# 12. Disaggregation by Location

Indicators are often disaggregated to study inequalities between groups.

Here we calculate SDG 5.3.1 separately for urban and rural areas. This is done exactly with the same code, but using the feature of  `group_by()` to get one result *by* category. 

*hint*: We removed (filter out) any observation where the area of residence is *"unknown"*.

```{r}
sdg_5_3_1_location <- wm_20_24 %>%
  filter(HH6 != "Unknown") %>%
  group_by( HH6 ) %>%
  summarise(
    denominator = sum(wmweight, na.rm = TRUE),
    numerator   = sum(wmweight[WAGEM < 18], na.rm = TRUE),
    SDG_5_3_1  = 100 * numerator / denominator
  )

# Making a nice table
kable(sdg_5_3_1_location,
      digits = 1, 
      caption = paste0("SDG 5.3.1 by Area - ",MyCountry," (", DataYear,")"))
```



# 13. Visualize Urban and Rural Differences

```{r}
sdg_5_3_1_location %>%
  filter(HH6 != "Unknown") %>%
ggplot()+ 
  aes(x = factor(HH6), y = SDG_5_3_1, fill = HH6) +
  geom_col() +
  geom_text(
    aes(label = paste(round(SDG_5_3_1, 2), "%")),
    hjust = 1.5,
    color = "white" ) +
  labs(
    title = "SDG 5.3.1 by Location",
    x = "Location ",
    y = "SDG 5.3.1 (in %)", 
    caption = paste0("Report for ",MyCountry," (", DataYear, ")"),
  ) +
  coord_flip() +
  theme_minimal()
```

# 14. Export the Results

The final results are saved as a CSV file, exported directly to the *output* directory.

```{r }
results <- sdg_5_3_1 %>%
  mutate(across(where(is.numeric), round, 2)) %>%  # Only 2 digits
  mutate(
    country = MyCountry,
    year = DataYear,
    unit = "Percent"
    ) %>%  # Order the columns...
 relocate(country, year, numerator, denominator,indicator, unit)

# The name of the output file depends on "Country" and "year"
write_csv(results, paste0("../Output/SDG_5_3_1_report_",MyCountry,"-", DataYear,".csv"))
```

# Conclusion

This tutorial illustrated a complete workflow for computing an SDG indicator using household survey microdata in R.

The workflow included:

- importing SPSS data,
- exploring variables,
- producing descriptive statistics,
- visualizing distributions,
- filtering the target population,
- calculating weighted indicators,
- producing disaggregated estimates,
- and exporting the results.

The same general approach can be adapted to many other SDG indicators derived from survey data.




