---
title: "Computing SDG 4.2.2 using MICS Data in R"
author: "Sinovia Moonie, Statistician/Lecturer, Statistical Institute for Asia and the Pacific"
subtitle: '*A simple report*'
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    code_download: true
  pdf_document:
    toc: yes
  word_document:
    toc: no
always_allow_html: true  # <-- bypass errors caused by interactive HTML (if any)
---

```{r setup, include=FALSE}
# We define the options for the chunks globaly here

knitr::opts_chunk$set( message = FALSE,
                       warning = FALSE,
                       results =TRUE,
                       echo = knitr::is_html_output() # Code only in html!
                       )

```

```{r country}
# Since we may want to use other data set from other countries/years
# Let us define the country and year as a parameter
country <- "Lao"
DataYear <- 2023
# Years taken into account for the definition of the indicator
RefYear <-  DataYear -1 # The last school year recorded
RefAge <- 5

```

---

# Introduction
This tutorial demonstrates how to compute the **SDG Indicator 4.2.2: percentage of children in organized learning** (one year before the official primary entry age),  using microdata from the **Multiple Indicator Cluster Survey (MICS)** for `r country`  (`r DataYear`).


The definition of **organised learning** is the following: 

> An organized learning programme is one that consists of a coherent set or sequence of educational activities designed to achieve pre-determined learning outcomes or the accomplishment of a specific set of educational tasks.  E.g.  early childhood education prgogramme



For **`r country`**, the reference cohort is children who were **`r RefAge`** years old in September **`r RefYear`**, meaning they were born between **October `r (RefYear-RefAge)-1`** and **September `r RefYear-RefAge`**. The data source is the Multiple Indicator Cluster Survey (MICS) household roster (`hl.sav`).

The tutorial follows a workflow for computing this indicator:

 INSERT WOKFLOW HERE


# 1: Install the appropriate R packages 

We use here a package manager `pacman` (another package). Once installed and loaded, it can facilitate the use of a list of packages

```{r packages}
# We use pacman here 
# install.packages("pacman")  # <-- Run only once

library(pacman)
pacman::p_load(
  haven,
  tidyverse,
  kableExtra, 
  modelsummary
)
```


# 2. Import the Data

The full questionnaire dataset (`hl.sav`) is imported into R using `read_sav()`.

```{r load-data}
rawdf <- read_sav("../Data/hl.sav")
```


The raw dataset contains **`r nrow(rawdf)`** observations across **`r ncol(rawdf)`** variables, covering all household members recorded in the survey.



# 3. Explore the Dataset

The indicator is constructed from a small set of variables

## Key Variables Used

| Variable    | Description                                          |
|-------------|------------------------------------------------------|
| `HL4`       | Sex of household member (1 = Male, 2 = Female)       |
| `HL5M`      | Date of birth — month                                |
| `HL5Y`      | Date of birth — year                                 |
| `HL6`       | Age of household member                              |
| `ED4`       | Ever attended school or ECE                          |
| `ED9`       | Attended school or ECE in current year (2022–2023)   |
| `ED10A`     | Level of education in current year (ECE = 0)         |
| `HH6`       | Location or Area (rural , Urban)                     |
| `windex5`   |  Wealth quintile                                     |
| `hhweight`  | Household sample weight                              |



##  Variable selection
 > Let us select the variable of interest in a new, smaller data set with only the variables of interest. 
 

```{r select-vars}
df <- rawdf %>%
  select( HL4, HL5M, HL5Y, HL6, ED4, ED9, ED10A, HH6, windex5, hhweight)


```

We have now a new data set  `df` with still **`r nrow(df)`** observations but only  **`r ncol(df)`** variables! Let us use some fancy package to get an overview of the values of these `r ncol(df)` variables: *`r names(df)`*.  
 
```{r Summary_all }
# We can use the package modelsummary for a nice descriptive table  
datasummary_skim(df , type = "numeric" )
```


After selection, the working dataset has **`r nrow(df)` rows** and **`r ncol(df)` columns**.

> Some variables ar not numbers (`numeric`) but categories (`factor`), We should adjust their types to avoid issues in the computation.


```{r redefine-type}
# We use as_factor to transform variables with their right type

df <- df %>%
  mutate(HL4 = as_factor(HL4), 
         HH6 = as_factor(HH6))
```


```{r Summary_factor }
# We can use the package modelsummary for a nice descriptive table  
datasummary_skim(df ,type = "categorical" )
```

---

# 4. Compute SDG 4.2.2 

## Define the Target Cohort
There are several step for computing the population of interest:

 - First, we exclude records with missing or invalid birth-month codes ( variable `HLM5` is   98, 99) 
 - Second, children who were **`r RefAge`** years old in September **`r RefYear `** were born between **October `r (RefYear-RefAge)-1`** and **September `r RefYear-RefAge`**.

> We need to identify the variables that give us that information 

| Variable    | Description                                          |
|-------------|------------------------------------------------------|
| `HL5M`      | Date of birth — month                                |
| `HL5Y`      | Date of birth — year                                 |





```{r cohort-filter}

# let's define the schooling period for this country/year

Year_Start <-(RefYear-RefAge)-1 
Year_End <- RefYear-RefAge

# Filter the observations
Pop_Children <- df %>%
  filter(     
    # Remove missing/invalid birth dates
    !is.na(HL5Y),
    !is.na(HL5M),
    !(HL5M %in% c(98, 99))
    ) %>%
  filter(    
    # Selecting the target cohort
    (HL5Y == Year_Start & HL5M >= 10) |
      (HL5Y == Year_End & HL5M <= 9)
  )
```


The cohort filter yields **`r nrow(Pop_Children)`** children born in the target window.

## Derive Participation Indicators

A child is counted as participating in **organised learning** if they attended school  (`ED9`) the **current year** (`r Year_Start` - `r Year_End` )  at the ECE or Grade 1 level ( `ED10A` = 0 or 1 ) .

```{r Population}
hl_422 <- Pop_Children %>%
    # Attending ECE or Grade 1 in current school year 
     mutate(
    organized_learning = ifelse(ED9 == 1 & ED10A %in% c(0, 1), TRUE, FALSE)
  )
```

```{r InLearning-count}
Nb_InLearning <-  sum(hl_422$organized_learning, na.rm = TRUE)

```

> There are **`r Nb_InLearning `** children in organized learning within **`r nrow(Pop_Children)`**  in the target cohort population. 

## National Estimate

The indicator is computed as a weighted proportion. Weights reflect the survey design.

```{r national-estimate}
sdg_4_2_2 <- hl_422 %>%
  summarise(
    Numerator   = sum(hhweight * organized_learning, na.rm = TRUE),
    Denominator = sum(hhweight, na.rm = TRUE),
    SDG_4_2_2   = 100 * Numerator / Denominator
  )

# print(sdg_4_2_2)
```

> The **National SDG 4.2.2 estimate is `r round(sdg_4_2_2$SDG_4_2_2, 1)`%** of 5-year-olds participated in organised learning in `r RefYear` - `r DataYear`.


# 5. Disaggregation by Sex

Here we disaggregate the indicator by sex (1 = Male, 2 = Female) to identify any gender gap in early childhood participation.

```{r disagg-sex}
sdg_4_2_2_sex <- hl_422 %>%
  group_by(HL4) %>%
  summarise(
    Numerator   = sum(hhweight * organized_learning, na.rm = TRUE),
    Denominator = sum(hhweight, na.rm = TRUE),
    SDG_4_2_2   = round(100 * Numerator / Denominator, 1)
  )

```


## Visualizing 

A simple bar chat make the male–female gap immediately readable.

```{r viz-sex, eval=FALSE}
sdg_4_2_2_sex %>%
  ggplot(aes(x = HL4, y = SDG_4_2_2, fill = HL4)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = paste0(SDG_4_2_2, "%")),
            vjust = -0.5, size = 4.5) +
  scale_fill_manual(values = c("#1a7abf", "#e05c2a")) +
  scale_y_continuous(limits = c(0, 110), labels = scales::percent_format(scale = 1)) +
  labs(
    title    = "SDG 4.2.2 by Sex",
    subtitle = paste(country, "- School Year", RefYear, "-", DataYear),
    x = NULL, y = "Children in organized learning (%)"
  ) +
   coord_flip() +
  theme_minimal(base_size = 13)
```

# 6. Disaggregation by Location 

Here we disaggregate the indicator by type of residence to identify any disparity in the location of households.  

```{r disagg-Area}
sdg_4_2_2__location <- hl_422 %>%
  mutate(HH6_N = ifelse(HH6 == "URBAN", HH6, "RURAL"))%>%
  group_by(HH6_N) %>%
  summarise(
    Numerator   = sum(hhweight * organized_learning, na.rm = TRUE),
    Denominator = sum(hhweight, na.rm = TRUE),
    SDG_4_2_2   = round(100 * Numerator / Denominator, 1)
  ) 

```


## Visualizing 


```{r viz-location, eval=FALSE}
sdg_4_2_2__location %>%
  ggplot(aes(x =HH6_N, y = SDG_4_2_2, fill = HH6_N)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = paste0(SDG_4_2_2, "%")),
            vjust = -0.5, size = 4.5) +
  scale_y_continuous(limits = c(0, 110), labels = scales::percent_format(scale = 1)) +
  labs(
    title    = "SDG 4.2.2 by location",
    subtitle = paste(country, "- School Year", RefYear, "-", DataYear),
    x = NULL, y = "Children in organized learning (%)"
  ) +
   coord_flip() +
  theme_minimal(base_size = 13)
```


# 7. Disaggregation by Wealth levels

Here we disaggregate the indicator following the wealth levels. We use the quintile of wealth available in the variable `windex5`

```{r disagg-wealth}
sdg_4_2_2_Wealth <- hl_422  %>%
  group_by(windex5) %>%  
  summarise(
    Numerator = sum(hhweight * organized_learning, na.rm = TRUE),
    Denominator = sum(hhweight, na.rm = TRUE),
    SDG_4_2_2  = round(100 * Numerator / Denominator, 1)
  )

```


```{r viz-wealth, eval=FALSE}
sdg_4_2_2_Wealth %>%
  ggplot(aes(x = windex5, y = SDG_4_2_2, fill = windex5)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = paste0(SDG_4_2_2, "%")),
            vjust = -0.1, size = 4.5) +
  scale_y_continuous(limits = c(0, 110), labels = scales::percent_format(scale = 1)) +
  labs(
    title    = "SDG 4.2.2 by Wealth levels (quintile)",
    subtitle = paste(country, "- School Year", RefYear, "-", DataYear),
    x = NULL, y = "Children in organized learning (%)"
  ) +
   coord_flip() +
  theme_minimal(base_size = 13)
```


# 8. Export Results

Both summary tables are saved as CSV files for further reporting or dissemination.

```{r export}
write_csv(sdg_4_2_2,     "../output/sdg_4_2_2_results.csv")
write_csv(sdg_4_2_2_sex, "../output/sdg_4_2_2_results_by_sex.csv")
write_csv(sdg_4_2_2__location, "../output/sdg_4_2_2_results_by_location.csv")
write_csv(sdg_4_2_2_Wealth, "../output/sdg_4_2_2_results_by_wealth.csv")
```


---
*Report generated with R `r paste(R.version$major, R.version$minor, sep = ".")` · `r format(Sys.time(), "%Y-%m-%d %H:%M")`*
