---
title: "Computing SDG 4_2_2 Using MICS Data in R"
author: "Sinovia Moonie, Statistician/Lecturer, Statistical Institute for Asia and the Pacific"
subtitle: '*A simple report*'
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    code_download: true
  word_document:
    toc: no
always_allow_html: true  # <-- bypass errors caused by interactive HTML (if any)
---

```{r setup, include=FALSE}
# We define the options for the chunks globaly here

knitr::opts_chunk$set( message = FALSE,
                       warning = FALSE,
                       results =TRUE,
                       echo = knitr::is_html_output() # Code only in html!
                       )

```

```{r country}
# Since we may want to use other data set from other countries/years
# Let us define the country and year as a parameter
country <- "Lao"
RefYear <-  "2023"
RefAge <- 5

```

---

# Introduction
This tutorial demonstrates how to compute the **SDG Indicator 4.2.2: percentage of children in organized learning** (one year before the official primary entry age),  using microdata from the **Multiple Indicator Cluster Survey (MICS)** for `r country`  (`r RefYear`).


The definition of **organised learning** is the following: 

> An organized learning programme is one that consists of a coherent set or sequence of educational activities designed to achieve pre-determined learning outcomes or the accomplishment of a specific set of educational tasks.  E.g.  early childhood education prgogramme



For **`r country`**, the reference cohort is children who were **`r RefAge`** years old in September **`r RefYear-1`**, meaning they were born between **October `r RefYear-RefAge-1`** and **September `r RefYear-RefAge`**. The data source is the Multiple Indicator Cluster Survey (MICS) household roster (`hl.sav`).

---

# Setup

## Load Libraries

```{r libraries}
library(tidyverse)
library(haven)
library(dplyr)
```

---


# Data Loading and Initial Inspection

## Load Raw Data

```{r load-data}
rawdf <- read_sav("../Data/hl.sav")
```

```{r dims-raw}
cat("Rows   :", nrow(rawdf), "\n")
cat("Columns:", ncol(rawdf), "\n")
```

The raw dataset contains **`r nrow(rawdf)` observations** across **`r ncol(rawdf)` variables**, covering all household members recorded in the survey.

## Quick Structure Check

```{r structure}
str(rawdf)
```

---

# Variable Exploration

The indicator is constructed from a small set of roster and education variables. Below we profile each one to check coding, value labels, and missing data before any recoding.

## Key Variables Used

| Variable    | Description                                          |
|-------------|------------------------------------------------------|
| `HL4`       | Sex of household member (1 = Male, 2 = Female)       |
| `HL5M`      | Date of birth — month                                |
| `HL5Y`      | Date of birth — year                                 |
| `HL6`       | Age of household member                              |
| `ED4`       | Ever attended school or ECE                          |
| `ED9`       | Attended school or ECE in current year (2022–2023)   |
| `ED10A`     | Level of education in current year (ECE = 0)         |
| `ED15`      | Attended school or ECE in previous year (2021–2022)  |
| `ED16A`     | Level attended in previous year (ECE = 0)            |
| `hhweight`  | Household sample weight                              |

## Sex (`HL4`)

```{r explore-sex}
class(rawdf$HL4)
table(rawdf$HL4)
attr(rawdf$HL4, "labels")
```

## Age (`HL6`)

```{r explore-age}
class(rawdf$HL6)
summary(rawdf$HL6)
```

## Current-Year School Attendance (`ED9`)

```{r explore-ed9}
class(rawdf$ED9)
table(rawdf$ED9)
attr(rawdf$ED9, "labels")
```

## Previous-Year School Attendance (`ED15`)

```{r explore-ed15}
class(rawdf$ED15)
table(rawdf$ED15)
attr(rawdf$ED15, "labels")
```

## Household Weight (`hhweight`)

```{r explore-weight}
class(rawdf$hhweight)
summary(rawdf$hhweight)
```

---

# Data Preparation

## Select Variables of Interest

We narrow the dataset to the ten variables needed for the indicator computation.

```{r select-vars}
df <- rawdf %>%
  select(HL1, HL4, HL5M, HL5Y, HL6, ED4, ED9, ED10A, ED15, ED16A, hhweight)
```

```{r dims-df}
cat("Rows   :", nrow(df), "\n")
cat("Columns:", ncol(df), "\n")
```

After selection, the working dataset has **`r nrow(df)` rows** and **`r ncol(df)` columns**.

---

# SDG 4.2.2 Computation

## Define the Target Cohort

Children who were **5 years old in September 2022** were born between **October 2016 and September 2017**. We exclude records with missing or invalid birth-month codes (98, 99).

```{r cohort-filter}
children_5_sep2022 <- df %>%
  filter(
    !is.na(HL5Y),
    !is.na(HL5M),
    !(HL5M %in% c(98, 99)),
    (HL5Y == 2016 & HL5M >= 10) |
      (HL5Y == 2017 & HL5M <= 9)
  )
```

```{r dims-cohort}
cat("Rows   :", nrow(children_5_sep2022), "\n")
cat("Columns:", ncol(children_5_sep2022), "\n")
```

The cohort filter yields **`r nrow(children_5_sep2022)` children** born in the target window.

## Derive Participation Indicators

A child is counted as participating in organised learning if they attended ECE (level 0) or Grade 1 (level 1) in **either** the current year (2022–2023) **or** the previous year (2021–2022).

```{r indicator-derive}
hl_422 <- children_5_sep2022 %>%
  mutate(
    # Attending ECE or Grade 1 in current school year 2022-2023
    organized_learning_current = ED9 == 1 & ED10A %in% c(0, 1),

    # Attended ECE or Grade 1 in previous school year 2021-2022
    organized_learning_previous = ED15 == 1 & ED16A %in% c(0, 1),

    # Either year counts
    organized_learning = organized_learning_current | organized_learning_previous
  )
```

```{r dims-hl422}
cat("Rows   :", nrow(hl_422), "\n")
cat("Columns:", ncol(hl_422), "\n")
```

```{r raw-count}
cat("Unweighted count in organised learning:", sum(hl_422$organized_learning, na.rm = TRUE), "\n")
```

## National Estimate

The indicator is computed as a weighted proportion. Weights reflect the complex survey design.

```{r national-estimate}
sdg_4_2_2 <- hl_422 %>%
  summarise(
    Numerator   = sum(hhweight * organized_learning, na.rm = TRUE),
    Denominator = sum(hhweight, na.rm = TRUE),
    SDG_4_2_2   = 100 * Numerator / Denominator
  )

print(sdg_4_2_2)
```

> **National SDG 4.2.2 estimate: `r round(sdg_4_2_2$SDG_4_2_2, 1)`%** of 5-year-olds participated in organised learning in 2022–2023.

---

# Disaggregation by Sex

Equity analysis is central to SDG monitoring. Here we disaggregate the indicator by sex (1 = Male, 2 = Female) to identify any gender gap in early childhood participation.

```{r disagg-sex}
sdg_4_2_2_sex <- hl_422 %>%
  group_by(HL4) %>%
  summarise(
    Numerator   = sum(hhweight * organized_learning, na.rm = TRUE),
    Denominator = sum(hhweight, na.rm = TRUE),
    SDG_4_2_2   = round(100 * Numerator / Denominator, 1)
  ) %>%
  mutate(Sex = case_when(
    as.numeric(HL4) == 1 ~ "Male",
    as.numeric(HL4) == 2 ~ "Female",
    TRUE ~ "Unknown"
  ))

print(sdg_4_2_2_sex)
```

---

# Export Results

Both summary tables are saved as CSV files for further reporting or dissemination.

```{r export}
write_csv(sdg_4_2_2,     "sdg_4_2_2_results.csv")
write_csv(sdg_4_2_2_sex, "sdg_4_2_2_results_by_sex.csv")
cat("Files written: sdg_4_2_2_results.csv, sdg_4_2_2_results_by_sex.csv\n")
```

---

# Suggested Visualisations

The charts below illustrate different aspects of the indicator. Adapt colours and labels to your institutional style guide.

## 4.1 — National Headline with a Reference Line

A simple gauge or horizontal bar comparing the national estimate to a policy target (e.g. 80 % or 100 %) gives an immediate sense of the coverage gap.

```{r viz-national, eval=FALSE}
library(ggplot2)

sdg_val <- sdg_4_2_2$SDG_4_2_2

ggplot(data.frame(x = "Lao PDR", y = sdg_val),
       aes(x = x, y = y)) +
  geom_col(fill = "#1a7abf", width = 0.4) +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "grey40") +
  geom_text(aes(label = paste0(round(y, 1), "%")),
            vjust = -0.6, size = 5, fontface = "bold") +
  scale_y_continuous(limits = c(0, 110), labels = scales::percent_format(scale = 1)) +
  labs(
    title    = "SDG 4.2.2 — Participation in Organised Learning",
    subtitle = "Lao PDR · School Year 2022–2023",
    x = NULL, y = "Participation Rate (%)"
  ) +
  theme_minimal(base_size = 13)
```

## 4.2 — Gender Disaggregation Bar Chart

Side-by-side bars make the male–female gap immediately readable.

```{r viz-sex, eval=FALSE}
sdg_4_2_2_sex %>%
  ggplot(aes(x = Sex, y = SDG_4_2_2, fill = Sex)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = paste0(SDG_4_2_2, "%")),
            vjust = -0.5, size = 4.5) +
  scale_fill_manual(values = c("Male" = "#1a7abf", "Female" = "#e05c2a")) +
  scale_y_continuous(limits = c(0, 110), labels = scales::percent_format(scale = 1)) +
  labs(
    title    = "SDG 4.2.2 by Sex",
    subtitle = "Lao PDR · School Year 2022–2023",
    x = NULL, y = "Participation Rate (%)"
  ) +
  theme_minimal(base_size = 13)
```

## 4.3 — Current vs. Previous Year Contribution

Stacked or grouped bars showing how many children are captured exclusively via the current year, exclusively via the previous year, or via both, helps assess data quality and year-on-year enrolment patterns.

```{r viz-year-contribution, eval=FALSE}
hl_422 %>%
  mutate(
    source = case_when(
      organized_learning_current & !organized_learning_previous ~ "Current year only",
      !organized_learning_current & organized_learning_previous ~ "Previous year only",
      organized_learning_current & organized_learning_previous  ~ "Both years",
      TRUE ~ "Neither"
    )
  ) %>%
  count(source, wt = hhweight, name = "weighted_n") %>%
  ggplot(aes(x = reorder(source, -weighted_n), y = weighted_n, fill = source)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::comma(round(weighted_n))), vjust = -0.4, size = 3.8) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Source of Organised Learning Participation",
    subtitle = "Weighted counts by year of attendance",
    x = NULL, y = "Weighted Population"
  ) +
  theme_minimal(base_size = 13)
```

## 4.4 — Age Distribution of the Cohort

A histogram of birth years (or exact ages) verifies that the cohort filter captured the intended children and flags any edge cases.

```{r viz-age-dist, eval=FALSE}
hl_422 %>%
  mutate(birth_year = as.numeric(HL5Y)) %>%
  ggplot(aes(x = birth_year, weight = hhweight)) +
  geom_histogram(binwidth = 0.5, fill = "#4caf8c", colour = "white") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Weighted Birth-Year Distribution of Target Cohort",
    subtitle = "Expected peak: 2016–2017",
    x = "Birth Year", y = "Weighted Count"
  ) +
  theme_minimal(base_size = 13)
```

---

*Report generated with R `r paste(R.version$major, R.version$minor, sep = ".")` · `r format(Sys.time(), "%Y-%m-%d %H:%M")`*
