---
title: "A step-by-step Data Analysis"
subtitle: "Data Science for Official Statistics"
author: "Christophe Bontemps"
date:   "`r Sys.Date()`"
output:
  html_document: 
    df_print: paged
    always_allow_html: true
    toc: yes
    toc_float: true
    keep_md: no
    code_folding: show
    code_download: true
  word_document:
    #reference_docx: KAZ-Template.docx
    toc: no 
    number_sections: true
  pdf_document:
    toc: yes
    df_print: kable
 
always_allow_html: true  # <-- bypass errors caused by interactive HTML
---


# Motivation

Descriptive statistics are essential in official statistics.  
However, they are often insufficient to understand **relationships** between variables.

They are essential, but:
  - They hide heterogeneity
  - They cannot isolate multiple effects
  - They may lead to misleading policy conclusions

We illustrate this here in this reproducible example. 

## Settings and libraries

```{r setup, include=FALSE}
# My default setting for this document
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	cache = FALSE,
	comment = NA
)
```

#### Install packages (*only once!*) and load them.

```{r InstallPackages, include=FALSE}
# List of the packages needed
Mypackages <- c(
   "data.table",  "RColorBrewer",
  "GGally", "knitr", "papeR", "modelsummary", 
  "sjPlot", "gt", "gtsummary", "tidyverse"  )

# Install any packages that are not yet installed ;-)
Newpackages <- setdiff(Mypackages, rownames(installed.packages()))
install.packages(Newpackages)

```

####  Loading the packages needed for the analysis 

```{r LoadPackages, include=FALSE}
# Load the packages we need
# tip: load tidyverse last to avoid issues with summarize!

cat("Installing all packages" )
for (pkg in Mypackages) {
  library(pkg, character.only = TRUE)
  cat("✓", pkg, "\n")
}
```

### A word on packages 

**Tidy data management packages**

- tidyverse
- data.table

**Plotting packages**

- RColorBrewer
- GGally

**Tables and sumary statistics** 

- modelsummary
- sjPlot
- gt 
- gtsummary

**Nice presentation of results**

- knitr
- papeR


One can also define some settings for the graphics and colors

```{r setup_graphics, include=FALSE}
# Setting the theme for graphics
theme_set(theme_minimal() + theme(panel.grid = element_blank()))
```


```{r colors, echo = FALSE}

# My colors:
SIAP.color <- "#0385a8"

# I will use my preferred colors for sex all along the document
sex_colors <- c(
  "Men" = "coral",       
  "Women" = "#9B59B6"     
)

age_colors <- c(
  "Young" = "#F39C12" ,
  "Middle" = "#C0504D",
  "Old" = "#8da0cb"
)

# Other colors
Some_colors <-c(
    "Young" = "#66c2a5",      # Soft blue
    "Middle" = "#F39C12",     # Warm orange
    "old" = "#C0504D"         # Muted burgundy/red
)
# Then: 
 # - Points and lines (and any others)
 #        scale_color_manual(values = sex_colors)
 #  - Bar chart
 #        scale_fill_manual(values = sex_colors)
 
  
```


```{r seed, include=FALSE}
# We'll see why this is important later
set.seed(2512) # For reproducibility
```

# Data

```{r dataloading}
data <-read.csv("../data/Household.csv")
```

**Variables:**

- **income**:  Household income (continuous)
- **education**: Adjusted years of education of household head (continuous)
- **age**: Age of household head (continuous)
- **sex**: Sex of household head ("Man" or "Woman)
- **education_level**: Level of education (3 levels)
- **age_group**: Age classification ("Young" , "Middle", "Old")



```{r cleaning, echo=FALSE}
# Data  cleaning
data <- data %>%
  select(income, age,  education, sex, education_level) %>%
  mutate(age_group = case_when( age <20 ~ "Young",
                                age >= 20 & age < 40 ~"Middle", 
                                age >=40 ~"Old"
                                ), 
         age_group = factor(age_group, levels = c("Young", "Middle", "Old")),
         sex= as.factor(sex), 
         education_level = as.factor(education_level)
  )
```


> First one can have a look at the data set (raw data). Here we dislay the 6 first observations. 

```{r head}
head(data)
```
 
 > **Question**:   How many variables in the data set? 

---

# Analysis


**Question:**  

> What determines the `income` for households?

---

##  Descriptive statistics

We have both continuous and categorical variables:

```{r StatSummary}
datasummary_skim(data, 
                 type = "numeric", 
                 title = "Continuous variables")
```


```{r StatSummaryC}
datasummary_skim(data,
                 type = "categorical",
                 title = " Categorical (factors)")
```

### Stats by sex

```{r StatGroup}
data_by_sex <-data %>%
  group_by(sex) %>%
  summarise(
    N = n(),
    mean_income = round(mean(income), 1),
    min_inc = round(min(income), 1),
    max_inc = round(max(income), 1),
    mean_educ = round(mean(education), 1),
    mean_age = round(mean(age), 1),
    .groups = "drop"
  )

kable(data_by_sex)
```

### Average income by sex

```{r StatBySex}
data_by_sex %>%
  ggplot(aes(sex, mean_income, fill = sex)) +
  geom_col(width = 0.2) +
  scale_fill_manual(values = sex_colors) +
  labs(
    title = "Average income by sex",
    subtitle = "NB: This is not a good graphic!",
    y = "Mean income",
    x = NULL
  ) +
  coord_flip()+
  theme(legend.position = "none")


```

**Tempting interpretation:**  
> Men earn more than women.


### A better representation of that 
```{r}
ggplot(data, aes(x = sex, y = income, fill = sex)) +
  # geom_boxplot(width = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "grey70" ) +
  scale_fill_manual(values = sex_colors) +
  labs(
    title = "Income by Sex",
    subtitle = "Showing the variability",
    x = NULL,
    y = "Income"
  ) +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none")
```



## Scatter plot: income vs education

```{r basic-scatterplot}
data %>%
ggplot() +
  aes(education, income) +
  geom_point(color = "lightblue", alpha = 0.8) +
  labs(
    title = "Scatter plot of Income vs education",
    x = "Education (Adjusted years)",
    y = "Income"
  ) 
```

This suggests a positive relationship, but variability is large.

##  Adding more information: Mutlivariate analysis

Looking at 2x2 relationships

## Income, and age

```{r ScatterIncomeAge}
ggplot(data, aes(age, income)) +
  geom_point(alpha = 0.6, color = "darkgrey") +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Income and age",
    x = "Age",
    y = "Income",
    color = "Age"
  )
```

> The relationship is still hard to interpret clearly.

### Income and Education , by age groups


```{r ScatterIncomeEdu, warning=FALSE}
ggplot(data, aes(education, income, color = age_group)) +
  geom_point(alpha = 0.6, position = position_jitter(width = 0.3, height = 0)) +
  facet_wrap(~ age_group) +
 # scale_color_manual(values = age_colors) +
  labs(
    title = "Income and education",
    subtitle = "Color is age group",
    x = "Education (Adjusted years)",
    y = "Income",
    color = "Age"
  ) +
  theme(
    panel.border = element_rect(color = "grey70", fill = NA, linewidth = 0.5),
    # axis.line.y = element_line(color = "black"),
    # axis.ticks.y = element_line(color = "black"),
    panel.spacing.x = unit(2, "lines"),  # Horizontal spacing
    panel.spacing.y = unit(1, "lines"),   # Vertical spacing (if wrapped)
    legend.position = "bottom"
)
```

## Can we plot all interactions? 

```{r Scattermatrix}
# Scatterplot matrix
GGally::ggpairs(data, 
        columns = c("income", "age", "education"),
        aes(color = sex, alpha = 0.3),
        title = "Distribution and relationships between Income, Age, and Education \n (by sex)",
        switch = "both",   # labels bottom and left
        upper = list(continuous = "cor"),      # Correlations in upper triangle
        lower = list(continuous = "points"),   # Scatterplots in lower triangle
        diag = list(continuous = "densityDiag")) + # Density plots on diagonal
  scale_color_manual(values = sex_colors) +
  scale_fill_manual(values = sex_colors)
```



# Limitations of descriptive statistics

## Why this is still not enough?

- Visual inspection is difficult sometimes
- Effects are hard to quantify
- Multiple variables are difficult to handle simultaneously, the number of interactions is important. 

This motivates multivariate regression.

# Regression model

## A simple  Linear model

Let's just assess the relation between $y$= income and  $X$ = Education

```{r RegressionUni, echo=TRUE}
modelUni <- lm(income ~ education, data = data)
summary(modelUni)
```

```{r}
tbl_regression(modelUni)

```


```{r }
modelsummary(
  modelUni,
  statistic = "std.error",
  stars = TRUE,
  output = "html"
)
```

## Interpretation

- The intercept is estimated at **`r round(coef(modelUni)[1], 2)`**.
- A one-unit increase in education is associated with an average increase of **`r round(coef(modelUni)[2], 2)`** units of income.

## Model fit

- R-squared: **`r round(summary(modelUni)$r.squared, 3)`**
- Adjusted R-squared: **`r round(summary(modelUni)$adj.r.squared, 3)`**

## Linear model with all variables

```{r, echo=TRUE}
modelMulti <- lm(income ~ education + age + sex , data = data)
summary(modelMulti)
```


```{r}
modelsummary(
  modelMulti,
  statistic = "std.error",
  stars = TRUE,
  output = "html"
)
```

```{r}
tab_model(
  modelMulti,
  title = "Regression of Income (Y)  on education, age and sex (Xs)",
  show.se = TRUE,
  show.p = TRUE,
  dv.labels = "Income",
  file = NULL
)
```


This model:

- Quantifies the effect of education
- Controls for age and sex

```{r regression_line}
ggplot(data, aes(education, income)) +
  geom_point(alpha = 0.4, color = SIAP.color) +
  geom_smooth(method = "lm", se = FALSE, color = "grey") +
  # geom_abline(intercept = 600, color = "darkred", linewidth = 1.5) +  # For illustation
  labs(
    title = "Income and education",
    subtitle = "Regression line adjusted for age and sex",
    x = "Education (Adjusted years)",
    y = "Income"
  ) 
```

```{r}
# Create prediction grid from the fitted model
pred_data <- expand.grid(
  education = seq(min(data$education),
                         max(data$education),
                         length.out = 100),
  sex = unique(data$sex),
  age = mean(data$age)
)

pred_data$pred_income <- predict(modelMulti, newdata = pred_data)

ggplot(data, aes(education, income, color = sex)) +
  geom_point(alpha = 0.3) +
  geom_line(
    data = pred_data,
    aes(education, pred_income, color = sex),
    linewidth = 1 ) +
  scale_color_manual(values = sex_colors)+
  labs(
    title = "Income and education by sex",
    subtitle = "Same regression model, adjusted for age",
    x = "Education (Adjusted years)",
    y = "Income",
    color = "Sex" ) +
  theme_minimal()
```


# Key message

## Take-away for official statistics

- Descriptive statistics are indispensable
- But they are not sufficient for policy analysis
- Regression is a natural and necessary extension

> **Descriptive statistics describe.  
Regression explains.**

---

# Bonus: A note on *partial* analysis


## Income–education relationship by sex
If we do the analysis by sex (that is for women only vs for Men only), we have this: 

```{r}
ggplot(data, aes(education, income, color = sex)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = sex_colors)+
  labs(
    title = "Income–education relationship by sex",
    x = "Education (Adjusted years)",
    y = "Income",
    color = "Sex"
  ) 
```

We now see:

- Education increases income for both sexes
- The effect is **stronger for young women** !!

> But we are not comparing the exact same thing!

We are not analysing the **whole population**, we are just doing **2 different analysis**, one for each sex, and therefore not comparing men and women fairly!   



