---
title: "Exercises with Survey data"
author: "Christophe Bontemps (SIAP)"
date: "29-08-2024"
output:
  word_document:
    toc: no
  html_document:
    toc: yes
    code_folding: show
    df_print: paged
  pdf_document:
    toc: no
---

```{rsetup, include=FALSE}
# Set global chunk options to hide code
knitr::opts_chunk$set(echo = FALSE)
```

# Overview
This document contains both the instructions and some code to help practice your data analytics skills.  We'll provide hints  as well. 
The goal is to work with some basic elements needed when using survey data and computing specific variables and indicators. Consider this as a self-paced guided tour, follow me!  

First we need to load some useful packages (libraries). 

> These libraries must have been installed beforefand using the bottom-right panel of RStudio. Normally you have these libraies already installed.  

```{r include=FALSE}
library(tidyverse) 
```

### Here are the steps to achieve:

* First we load the data
* Second we compute some averages 
* Third we define new variables 

> For this exercise, you will not need to download data on your computer. We read the data directly in R, from a server, and then store that in a data frame in R.


# Prerequisite: 

* Save the file ***ExerciseWithSurveyData.Rmd*** on a folder in your computer and open it with RStudio (hint: double-click) 
*  Change the author with your name (*hint: rpelace my name with yours and be careful of the keeping the " "* )
* Compile or `knit` the document to create an Html or Word file

**All along this exercise, answer all questions in the .Rmd document in RStudio**  

# Working with Survey data 

###  First we load the data
The data we will be working with are survey-type data, 

```{r }
household_survey <- read.csv2("..Data/Household_Survey.csv")
```

Now we have a data set, we can view it on the top-right panel. 

## Exercise 1:

*Write your answers directly below, or create an R code chunk to compute the solution*. 

 - How many observations (households) has this file? 
- How many variables? 

*(hint: you can use the function `nrow()`)*

**Your answer:**


## Exercise 2:
What is the format of the variable *location* ?  (hint: Look in the *environment tab* in the top-right panel and see the description of the variable *location* in the data frame *household_survey*. Alternatively, you can use the function `str()`  in a code chunk. 

**Your answer:**



## Exercise 3:
Explain what this code is achieving. 


```{r}
# Code using filter() function
household_rural <- household_survey %>%
  filter(location =="Rural" )
```

Hint: The code above is doing exactly the same thing (but storing inwith a different name) than the code below: 

```{r}
## NB This can be done with subset() too
household_rural_bis <- subset(household_survey, location =="Rural" )
```

**Your answer:**






# Computing new variables 

We want to compute the average income of the rural population: We do that in 4 steps:

```{r}
# Step 1: Create a data frame with only the  rural population (using filter()) 
household_rural <- household_survey %>%
  filter(location =="Rural" )

# Step 2: Calculate total income for the rural population
total_rural  <- household_rural %>%
  summarize(total = sum(income))

# Step 3: Calculate the number of rural households 
Nb_rural <-  household_rural %>%
  count()

# Step 4: Compute the average income for the rural population
avg_income_rural <- total_rural$total / Nb_rural$n

# Display the result
print(avg_income_rural)

```
So, the average income for **rural** population  is *`r round(avg_income_rural)`*. 

## Exercise 4

Replicate this exercise  for the *Urban* population (hint: copy paste the previous code). Name the result *avg_income_urban* and compare with rural population. Which one is greater?
 
**Your answer:**




# Working with population in surveys

### But Wait! This file is a survey, and We have to take into account the weight of each observation! 

## Exercise 5

After examining the data frame  *household_survey*  identify the variable that represents the weight of each household. (hint: This variable must be numerical, with values between 0 and 1)

```{r}
summary(household_survey$weight)
```

**Your answer** (Hint, you can also use the function `summary()`in an R chunk)  




**Extra bonus question**
Is there something strange with these weights? 

***Yes!*** Having a household with weight equal to 0 is not right, this would mean that we are not counting that household at all in our computation


## Computing the weighted average for rural population 
 We need to change our way to compute the sums in our previous computation
 
## Exercise 6 

Create a new variable *total_weighted_rural* in the data.frame *household_rural*  that will multiply the  income of each household (each line) with its weight. Hint: you can use the `%>%` operator and the `mutate()` function to create a new variable.

**Your answer**  




Below is the code that computes the weighted average income for the rural population. Please check the difference and how we have modified the 3 steps from previous average computation.   
 
```{r}
# Step 2bis: Calculate total weighted  income
total_weighted_rural  <- household_rural %>%
  mutate(weighted_income = income * weight) %>%   # <- This computes the weighted income
  summarize(total_weighted_income = sum(weighted_income))


# Step 3bis: Calculate the weighted number of households 
Nb_weighted_rural <-  household_rural %>%
  summarize(total_weights = sum(weight))
  
# Step 4bis: Compute the weighted average income
avg_weighted_income <- total_weighted_rural$total_weighted_income / Nb_weighted_rural$total_weights

print(avg_weighted_income)

```
## Exercise 7  

Compare the weighted average income (*avg_weighted_rural* ) with the simple average for the rurla population. Which one is greater?    

**Your answer**  




> Note that we can also achieve the same computation in a few line of code ;-)

```{r}
avg_weighted_income2 <- household_survey %>%
   filter(location =="Rural" ) %>%            # Keeping only rural households
   summarize(weighted_avg_income = sum(income * weight) / sum(weight))   # Computing weighted average

# printing the result (stored in a data frame)
print(avg_weighted_income2$weighted_avg_income)
```

# Creating a specific category 

> Task: Create a new column (variable)  *income_category* that categorizes households into "Low",  or "High" income based on the value of income. Use the following thresholds:

* "Low" for income below $50,000
* "High" for income above $70,000

Here is the code that uses the `ifelse()`function for testing whether a condition is true ( whether *income* is lower than 5000, here ) and returns "low" if yes and "high if not in this case,  while the  `mutate()` function creates a new variable *income_category* 

```{r}
household_survey <- household_survey %>%
  mutate(income_category = ifelse(income < 50000, "Low", "High"))
```

## Exercise 8: 

> Task: Create a new column *special_status* that indicates whether a household should receive special consideration based on two conditions:

* The household_size is greater than 5
AND
*the income is below $50,000.

Hint: Use the previous code and change the simple condition for the following condition: 
`household_size > 5 & income < 50000` as first argument of the `ifelse()` function.  The `&` symbol stands for AND between two logical conditions. 

**Your answer**  





```{r}
# Solution here 
household_survey <- household_survey %>%
  mutate(special_status = ifelse(household_size > 5 & income < 50000, 
                                  "Special Consideration", 
                                  "No Special Consideration"))
```

# Congratulations! 

This is the end. Normally you should be able to *knit* this file and send the output (word format) to *christophe.bontemps@un.org.*

Thanks! 