---
title: "Exercises with Survey data"
author: "Christophe Bontemps (SIAP)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  word_document:
    toc: no
  html_document:
    toc: yes
    code_folding: show
    df_print: paged

header-includes:
  - \usepackage{float}
  - \floatplacement{table}{H}

---


```{r setup, include=FALSE}
# My default setting for this document
knitr::opts_chunk$set(
  echo = FALSE, 
	message = FALSE,
	warning = FALSE,
	cache = FALSE,
	comment = NA
)
```



# Overview
This document contains both the instructions and some code to help practice your data analytics skills.  We'll provide hints  as well. 
The goal is to work with some basic elements needed when using survey data and computing specific variables and indicators. Consider this as a self-paced guided tour, follow me!  

First we need to load some useful packages (libraries). 

> These libraries must have been installed beforehand using the bottom-right panel of RStudio. Normally you have these libraries already installed.  

```{r , include=FALSE}
library(tidyverse) 
library(modelsummary)
library(knitr)
```



```{r datageneration, include=FALSE}
# set.seed(42)
# n <- 2000
# 
# householdsize <- sample(1:8, n, replace = TRUE,
#                          prob = c(0.10,0.20,0.25,0.20,0.12,0.07,0.04,0.02))
# gender    <- sample(c("Male","Female"), n, replace = TRUE, prob = c(0.60,0.40))
# location  <- sample(c("Urban","Rural","Suburb"), n, replace = TRUE,
#                          prob = c(0.45, 0.35, 0.20))
# 
# # Age strongly tied to location → collinearity (so location absorbs age)
# loc_mean_age <- ifelse(location == "Rural", 55,
#                        ifelse(location == "Suburb", 45, 33))
# age <- pmax(18, pmin(85, round(rnorm(n, mean = loc_mean_age, sd = 5))))
# # --- Income  ---
# log_income <- (
#   9.0
#   + 0.35 * (gender == "Male")
#   + 0.45 * (location == "Urban")
#   - 0.30 * (location == "Rural")
#   + 0.04 * householdsize
#   + rnorm(n, 0, 0.60)
# )
# income <- round(exp(log_income), 2)
# 
# # --- Weights ----
# weight <- round(runif(n, 0.10, 1.00), 4)
# zero_idx            <- sample(n, 15)
# weight[zero_idx]    <- 0
# outlier_idx         <- sample(setdiff(seq_len(n), zero_idx), 2)
# weight[outlier_idx] <- 99
# 
# # ---- Export ----
# survey <- data.frame(householdsize, income,
#                      gender, age, location, weight)
# write.csv(survey, "../data/Household2026.csv", row.names = FALSE)
```



### Here are the steps to achieve:

* First, load the data
* Second, create new variables
* Third, define new variables 
* Finally, run a simple regression model


# Prerequisite: 
* Check that you have a folder structure, with directories named `code`, `data`, `output` (and other if you want). The data set *Household_Survey.csv*  must be in the `data` folder, while the file *ExerciseWithSurveyData.Rmd*  should be in the `code` folder. 

* Open the file *ExerciseWithSurveyData.Rmd* with RStudio 
* Change the author with your name (*hint: be careful of  keeping  " "* )


> All along this exercise, you must answer all questions in the .Rmd document in RStudio only 


# Working with Survey data 

###  First, load the data
The data are survey-type data. The file is a csv file named `Household2026.csv` in the folder `data`. Use the function `read.csv()` to load the data. 

```{r }
household_survey <- read.csv("../data/Household2026.csv")
```

Now we have a data set, you can view it (use the  top-right panel or `summary()`). 


## Exercise 1:

*Write your answers directly below, or create an R code chunk to compute the solution*. 

- How many observations (households) has this file? 
- How many variables? 

*(hint: you can use the function `nrow()`)*

**Your answer:**



You should see the following variables:  *`r names(household_survey)`*



**Alternatively**, you can use the code below: 

```{r echo=TRUE}
datasummary_skim(household_survey, 
                 title = "Continuous variables")
```


## Exercise 2:

What is the format of the variable `location` ? 

**Hint:** Look in the *environment tab* in the top-right panel and see the description of the variable *location* in the data frame *household_survey*. 

Alternatively, you can use the function `str()`  in a code chunk. 

**Your answer:**



## Exercise 3:
Explain what this code is achieving. (run the code) 


```{r ruraldata, echo = TRUE}
# Code using filter() function
household_rural <- household_survey %>%
  filter(location =="Rural" )
```


**Your answer:**






# Computing new variables 

We want to compute the average income of the rural population: We do that in 4 steps:

```{r rural, echo =TRUE}
# Step 1: Create a data frame with only the  rural population (using filter()) 
household_rural <- household_survey %>%
  filter(location =="Rural" )

# Step 2: Calculate total income for the rural population
total_rural  <- household_rural %>%
  summarize(total = sum(income))

# Step 3: Calculate the number of rural households 
Nb_rural <-  household_rural %>%
  count() %>%
  pull()

# Step 4: Compute the average income for the rural population
avg_income_rural <- total_rural$total / Nb_rural



```

So, the average income for **rural** population  is *`r round(avg_income_rural, 2)`*. 

## Exercise 4

Replicate this exercise  for the *Urban* population. Name the result *avg_income_urban* and compare with rural population.

- Which one is greater?


 
**Your answer:**




# Working with population in surveys

### But Wait! This file is a survey, and We have to take into account the weight of each observation! 

## Exercise 5

After examining the data frame  *household_survey* 
-Identify the variable that represents the weight of each household. 

**Hint:** This variable is numerical, with values between 0 and 1.

- provide some summary statistics of this variable (you can  use the function `summary()`)  

```{r , include= FALSE}
summary(household_survey$weight)
```

**Your answer** 




**Extra bonus question**
Is there something important to notice with these weights? 




## Computing the weighted average for rural population 
We need to change our way to compute the sums in our previous computation
 
## Exercise 6 

Create a new variable *total_weighted_rural* in the data.frame *household_rural*  that will multiply the income of each household with its weight. 

**Hint:** you can use the `%>%` operator and the `mutate()` function to create a new variable.

**Your answer:**  




Below is the code that computes the weighted average income for the rural population. Please check the difference and how we have modified the 3 steps from previous average computation. 

 
```{r, echo = TRUE}
# Step 2bis: Calculate total weighted  income
total_rural_w  <- household_rural %>%
  mutate(income_w = income * weight) %>%   # <- This computes the weighted income
  summarize(total_w = sum(income_w))


# Step 3bis: Calculate the weighted number of households 
Nb_rural_w <-  household_rural %>%
  summarize(total_w = sum(weight)) %>%
  pull()
  
# Step 4bis: Compute the weighted average income
avg_income_rural_w <- total_rural_w$total_w / Nb_rural_w

print(avg_income_rural_w)

```
## Exercise 7  

Compare the weighted average income (*avg_weighted_rural* ) with the simple average for the rural population. Which one is greater?    

**Your answer:**  




> Note that we can also achieve the same computation in a few line of code ;-)

```{r, echo=TRUE}
avg_income_rural_w2 <- household_survey %>%
   filter(location =="Rural" ) %>%            # Keeping only rural households
   summarize(avg_income_w =
               sum(income * weight) / sum(weight)) %>%  # Computing weighted average
  pull() 

# printing the result (stored in a data frame)
print(avg_income_rural_w)
```

# Creating a specific category 

> Task: Create a new column (variable)  *income_category* that categorizes households into "Low",  or "High" income based on the value of income. Use the following thresholds:

* "Low" for income below $50,000
* "High" for income above $70,000

Here is the code that uses the `ifelse()`function for testing whether a condition is true ( whether *income* is lower than 5000, here ) and returns "low" if yes and "high if not in this case,  while the  `mutate()` function creates a new variable *income_category* 

```{r}
household_survey <- household_survey %>%
  mutate(incomecategory = ifelse(income < 50000, "Low", "High"))
```

## Exercise 8: 

> Task: Create a new column *special_status* that indicates whether a household should receive special consideration based on two conditions:

* The household size is greater than 5
AND
* the household income is below $50,000.

Hint: Use the previous code and change the simple condition for the following condition: 
`householdsize > 5 & income < 50000` as first argument of the `ifelse()` function.  The `&` symbol stands for AND between two logical conditions. 

**Your answer:**  





```{r}
# Solution here 
household_survey <- household_survey %>%
  mutate(special_status = ifelse(householdsize > 5 & income < 50000, 
                                  "Special Consideration", 
                                  "No Special Consideration"))
```


# Run and analyse a simple regression

For this exercise, we will **not** take weights into consideration (although this is obviously possible). The function used for running a regression is `lm()`

## Exercise 9

Regress `income` on the following explanatory variables: 

- `householdsize`
- `gender`
- `location`
- `age`

The regression result should similar to the table below: 

**Hint:** proceed in two steps: 

- Compute the regression using the function `lm()` and store the result in an object
- Use the function summary() to get the results. 



```{r }
reg <- lm(income~householdsize+gender+location+age, data = household_survey)

# table of results
modelsummary(reg, stars = TRUE,
shape = term ~ model + statistic,
statistic = c("{std.error}", "{statistic}", "{p.value}"),
gof_map = c("nobs", "r.squared"),
fmt = 2,
escape = TRUE
)

```

# Exercise 10:

> Interpret the results 

- Which parameters/variables are significant? 
- What can you say about the result of the regression? 
- How do you interpret the coefficient associated with `genderMale`


**Your answer:**  



# Congratulations! 

This is the end. Normally you should be able to *knit* this file and send the output (word or Html  format) to *christophe.bontemps@un.org* by Friday 3 July. 

Thanks! 