---
title: "Exercises with Survey data"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    toc: yes
    toc_float: true
    keep_md: no
    code_folding: show
    code_download: true
    df_print: paged
  word_document:
    toc: no
header-includes:
  - \usepackage{float}
  - \floatplacement{table}{H}
---


```{r setup, include=FALSE}
# My default setting for this document
knitr::opts_chunk$set(
  echo = FALSE, 
	message = FALSE,
	warning = FALSE,
	cache = FALSE,
	comment = NA
)
# avoid scientific notation
options(scipen = 999)

```



# Overview
This document contains both the instructions and *some* code to help practice your data analytics skills. It is aligned with what we have see during the lectures and, don't worry, I'll provide hints as well ;-). 
The goal is to work with some basic elements needed when using survey data and computing specific variables and indicators. 

## Instructions: 
Follow the steps, and **answer directly in the Rmd document!**.  One completed, you should be able to `knit` the document and send me and html file. 

> *This exercise should not take you more than one hour*


# Let's start!

First we need to load some useful packages (libraries). 

> These libraries must have been installed beforehand using the bottom-right panel of RStudio. Normally you have these libraries already installed.  

```{r , include=FALSE}
library(tidyverse) 
library(modelsummary)
```


## Here are the steps to achieve:

* First, load the data
* Second, create new variables
* Third, define new variables 
* Finally, run a simple regression model


## Prerequisite: 
Check that you have a folder structure, with directories named: 

- `code`,
- `data`,
- `output` (and other if you want).

> The data set *Household2026.csv*  **must be** in the `data` folder, while the file *ExerciseWithSurveyData.Rmd*  should be in the `code` folder. 

* Open the file *ExerciseWithSurveyData.Rmd* with RStudio 
* Change the author with your name (*hint: be careful of  keeping  " "* )


> **All along this exercise, you must answer all questions in the .Rmd document in RStudio only** 


# Working with Survey data 

###  First, load the data
The data are survey-type data. The file is a csv file named `Household2026.csv` in the folder `data`. Use the function `read.csv()` to load the data. 



```{r }
# fill the blanks
household_survey <- read.csv(" ..... ")
```

Now we have a data set, you can view it (use the  top-right panel or `summary()`). 


## Exercise 1:

*Write your answers directly below, or create an R code chunk to compute the solution*. 

- How many observations (households) has this file? 
- How many variables? 

*(hint: you can use the functions `nrow()` or `ncol()`)*

**Your answer:**





You should see the following variables:  *`r names(household_survey)`*


**Alternatively**, you can use the code below: 

```{r echo=TRUE}
datasummary_skim(household_survey, 
                 title = "Descriptive statistics")
```


## Exercise 2:

What is the format of the variable `location` ? 

**Hint:** Look in the *environment tab* in the top-right panel and see the description of the variable *location* in the data frame *household_survey*. 

Alternatively, you can use the function `str()`  in a code chunk. 

**Your answer:**



## Exercise 3:
Explain what this code is achieving. (run the code) 


```{r ruraldata, echo = TRUE}
# Code using filter() function
household_rural <- household_survey %>%
  filter(location =="Rural" )
```


**Your answer:**






# Computing new variables 

We want to compute the average income of the rural population: We do that in 4 steps:

```{r rural, echo =TRUE}
# Step 1: Create a data frame with only the  rural population (using filter()) 
household_rural <- household_survey %>%
  filter(location =="Rural" )

# Step 2: Calculate total income for the rural population
total_rural  <- household_rural %>%
  summarize(total = sum(income))

# Step 3: Calculate the number of rural households 
Nb_rural <-  household_rural %>%
 nrow()

# Step 4: Compute the average income for the rural population
avg_income_rural <- total_rural$total / Nb_rural


```



## Exercise 4:

Replicate this exercise  for the *Urban* population. Name the result *avg_income_urban* and compare with rural population.

- Which one is greater?


```{r}
# Write your code here

```

 

 
**Your answer:**




# Working with population in surveys

### But Wait! This file is a survey, and We have to take into account the weight of each observation! 

## Exercise 5:

After examining the data frame  *household_survey* , identify the variable that represents the weights for each household. 


**Hint:** This variable is numerical, with values between 0 and 1.

> Provide some summary statistics of this variable (you can  use the function `summary()`)  



**Your answer** 

```{r}
# Compute something that could help you describe the weights

```




**Extra bonus question**
Is there something important to notice with these weights? 




## Computing the weighted average for rural population 
We need to change our way to compute the sums in our previous computation
 
## Exercise 6: 

Within the data.frame *household_rural* , create a new variable *total_weighted_rural*  that will multiply the income of each household with its weight. 

**Hint:** you can use the `%>%` operator and the `mutate()` function to create a new variable.

**Your answer:**  


## Exercise 7:  

Below is the code that computes the weighted average income for the rural population. Please check the difference and how we have modified the 3 steps from previous average computation. 

 
```{r, echo = TRUE}
# Step 2bis: Calculate total weighted  income
total_rural_w  <- household_rural %>%
  mutate(income_w = income * weight) %>%   # <- This computes the weighted income
  summarize(total_w = sum(income_w))


# Step 3bis: Calculate the weighted number of households 
Nb_rural_w <-  household_rural %>%
  summarize(total_w = sum(weight)) %>%
  pull()    # <- this function helps extracting a single value 
  
# Step 4bis: Compute the weighted average income
avg_income_rural_w <- total_rural_w$total_w / Nb_rural_w

print(avg_income_rural_w)

```
  

Compare the weighted average income (*avg_weighted_rural* ) with the simple average for the **rural** population. Which one is greater?    

**Your answer:**  




> Note that we can also achieve the same computation in a few line of code ;-)

```{r, echo=TRUE}
avg_income_rural_w2 <- household_survey %>%
   filter(location =="Rural" ) %>%            # Keeping only rural households
   summarize(avg_income_w =
               sum(income * weight) / sum(weight)) %>%  # Computing weighted average
  pull() 

# printing the result (stored in a data frame)
print(avg_income_rural_w)
```

# Creating a specific category 

> Task: Create a new column (variable)  *income_category* that categorizes households into "Low",  or "High" income based on the value of income. Use the following thresholds:

* "Low" for income below $50,000
* "High" for income above $70,000

Here is the code that uses the `ifelse()`function for testing whether a condition is true ( whether *income* is lower than 5000, here ) and returns "low" if yes and "high if not in this case,  while the  `mutate()` function creates a new variable *income_category* 

```{r, echo=TRUE}
household_survey <- household_survey %>%
  mutate(incomecategory = ifelse(income < 50000, "Low", "High"))
```

## Exercise 8: 

> Task: Create a new column *special_status* that indicates whether a household should receive special consideration based on two conditions:

* The household size is greater than 5, **and**
* the household income is below $50,000.

**Hint:** Use the previous code and change the simple condition for the following condition: 
`householdsize > 5 & income < 50000` as first argument of the `ifelse()` function.  The `&` symbol stands for AND between two logical conditions. 

**Your answer:**  Compute the number of households under the *special status* category. 






# Run and analyse a simple regression

For this exercise, we will **not** take weights into consideration (although this is obviously possible). The function used for running a regression is `lm()`.

## Exercise 9:

Regress `income` on the following explanatory variables: 

- `householdsize`
- `gender`
- `location`
- `age`



```{r }


```

**Hint:** proceed in two steps: 

- Compute the regression using the function `lm()` and store the result in an object
- Use the function summary() to get the results. 



# Exercise 10:

> Interpret the results 

- Which parameters/variables are significant? 
- What can you say about the result of the regression? 
- How do you interpret the coefficient associated with `genderMale`?


**Your answer:**  



# Congratulations! 

This is the end. Normally you should be able to *knit* this file and send the output (word or Htmlformat) to *christophe.bontemps@un.org* by Friday 3 July. 

Thanks! 