---
title: "A Fully Automated Report- revised during the class"
subtitle: "TAPOS" 
author: "Christophe Bontemps"
date: "`r format(Sys.time(), '%A, %d %B %Y, %H:%M')`"
output:
  html_document:
    toc: yes
    code_folding: hide
    code_download: true 
    toc_float: true
    df_print: paged
  pdf_document:
    toc: no
  word_document:
    toc: yes
---

```{r, include=FALSE}
# Select one product for the detailed analysis
# Other options: "Wood", "Coffee", "Alcoholic Drinks"
Mygood <- "Wood"  

```

# An automatic report on  ***`r Mygood`*** 

```{r setup, include=FALSE}
knitr::opts_chunk$set( message = FALSE, warning = FALSE, results =TRUE, echo = TRUE) 
```

**Loading libraries**


```{r loadinglibraires, echo = FALSE}
# Load the required libraries

# For manipulating data
library(tidyverse)

# For tables display
library(flextable)
```

This is a report using **Rmarkdown** with a very simple structure.

```{r test1, fig.height=3, fig.width=2, message=TRUE, warning=TRUE}

```


* First, we load the data
* Second, we analyse the whole data set
* Finally, we focus on a specific good. Here we will focus on ***`r Mygood`*** 

```{r TradeDataLoading}
# We can use the data locally or from SIAP server

# Mydata <- read.csv("Data/TradeData.csv")
 Mydata <- read.csv("https://www.unsiap.or.jp/on_line/2024/RAP/Data/TradeData.csv")

```

## Raw Data Analysis - all products 
I have **here  `r  nrow(Mydata)`  observations**

```{r}
nrow(Mydata)
```

## Summary statistics

```{r }
summary(Mydata$Stat_Value)
```

# Data Manipulation

## Defining the variables with the right "type" 

R is a ***typed*** language with six primary data *types* that variables can hold (numeric,integer, character,  logical, ...). The type determines what kind of data the variable stores and which operations can be performed on it. 

> It is important to work with a data frame where each variable has the right type

```{r}
# Right type
Mydata <- Mydata %>%
  mutate(Year = as.integer(Year), 
         Stat_Value = as.numeric(Stat_Value)
  )
```



```{r CommodityByType}
# Calculate total value by principle export type
CommodityByType <- Mydata %>%
  group_by(Principle_Exports) %>%
  summarise(total = sum(Stat_Value, na.rm = TRUE)) %>%
  arrange(desc(total))

```

Here this are the principle exports by type

```{r TableCommodityByType}
# Table display
# To reduce the homogenization efforts and the number of functions to be called, 
# it is recommended to define formatting properties once and for all

set_flextable_defaults(
  font.size = 10,   table.layout = "autofit")

# Printing the table
flextable(CommodityByType)
```

# Graphical Analysis

Principal Exports over the period

```{r}
# Plot the exports
p <-CommodityByType  %>%
  ggplot()+
    aes(x = reorder(Principle_Exports, total), y = (total)) +
  geom_bar(stat = "identity", 
             fill= "lightsteelblue1") +
  labs(title ="Exports from Esdeegia [ 2000-2010] ", 
       subtitle = "Esdeegia National Statistical Office", 
       x= "Exports categories",
       y = "Vatu") +
  coord_flip() +
  theme_minimal()
p
```


# Focus on `r Mygood` exports

We have `r nrow(filter(Mydata, Principle_Exports== Mygood))` observations for the product `r Mygood`


```{r}
Mydata %>%
  filter(Principle_Exports == Mygood) %>%
  group_by(Year) %>%
  summarise(total = sum(Stat_Value, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = total)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_x_continuous(breaks = scales::breaks_pretty()) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    title = paste0("Exports of ", Mygood , " over Time"),
    subtitle = "Period [2000-2010]", 
    caption =  "Esdeegia National Statistical Office",
    x = "Year",
    y = "Total Export Value"
  ) +
  theme_minimal()
```


```{r Alluvial}
library(ggalluvial)

p<- Mydata %>%
  filter(Principle_Exports == Mygood) %>%
  filter(EXPORT_COUNTRY != IMPORT_COUNTRY) %>%        # remove same country flows
  group_by(EXPORT_COUNTRY, IMPORT_COUNTRY) %>%
  summarise(total = sum(Stat_Value, na.rm = TRUE), .groups = "drop") %>%
  mutate(EXPORT_COUNTRY = fct_reorder(EXPORT_COUNTRY, total, sum)) %>%  # sort by total
  ggplot(aes(
    axis1 = EXPORT_COUNTRY,
    axis2 = IMPORT_COUNTRY,
    y = total
  )) +
  geom_alluvium(aes(fill = EXPORT_COUNTRY), alpha = 0.7) +
  geom_stratum(color = NA, fill = NA) +           # removes boxes
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3) +
  scale_x_discrete(limits = c("Export Country", "Import Country"),
                   expand = c(0.1, 0.1)) +
  labs(
    title = paste("Trade Flow:",Mygood," Exports"),
    y = "Total Export Value",
    fill = "Export Country"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p

```

```{r, include=FALSE}
# Word document cannot handle interactive documents
# uncomment if output is word
 knitr::knit_exit()
```

# Interactive version

```{r}
library(plotly)
ggplotly(p)
```


```{r}
library(networkD3)
library(tidyverse)

# Prepare data
flows <- Mydata %>%
  filter(Principle_Exports == Mygood) %>%
  filter(EXPORT_COUNTRY != IMPORT_COUNTRY) %>%        # remove same country flows
  group_by(EXPORT_COUNTRY, IMPORT_COUNTRY) %>%
  summarise(total = sum(Stat_Value, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total))

# networkD3 needs nodes and links as separate objects
nodes <- data.frame(
  name = unique(c(flows$EXPORT_COUNTRY, flows$IMPORT_COUNTRY))
)

links <- flows %>%
  mutate(
    source = match(EXPORT_COUNTRY, nodes$name) - 1,  # 0-indexed
    target = match(IMPORT_COUNTRY, nodes$name) - 1
  ) %>%
  rename(value = total) %>%
  select(source, target, value)

# Plot
sankeyNetwork(
  Links = links,
  Nodes = nodes,
  Source = "source",
  Target = "target",
  Value = "value",
  NodeID = "name",
  fontSize = 12,
  nodeWidth = 20,
  sinksRight = TRUE
)
```

```{r}
knitr::knit_exit()
```


