---
title: "Analysing Child Marriage with Logistic Regression"
subtitle: "A step-by-step case study with Lao data "
author: "Christophe Bontemps SIAP"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document: default
  html_document:
    code_folding: show
    highlight: tango
    number_sections: no
    theme: lumen
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( message = FALSE, warning = FALSE, results =TRUE, echo = TRUE) 
```


### Good practices and file management {-}

We need a minimum of organisation in the data and code folders, as well as some R packages.
The folder is organized  with:

- Data  <- *Original (raw) data will be there*
- Outputs <- *Any result will go there*
- Documentation  <- *Docs and explanations, additional resources*

In our organization, the code (.R and .Rmd files) will be at the root of the directory. In general, it is recommended to have a directory with `code`. 

We also need some packages: 

```{r packages}
# Importing SPSS file
library(haven)

# Tidy data management packages
library(tidyverse)
library(data.table)

# Plotting packages
library(ggplot2)
library(RColorBrewer)

# Nice presentation of results
library(knitr)
library(papeR)
library(visreg)
library(ggpubr)

# Tables 
library(modelsummary)
library(gt) 
library(gtsummary)

```

#  Understanding child marriage

## Read the data

Our data set being in the directory `Data`, we need to specify the right path when loading the data. 

```{r}
wm <- read_spss(file ="Data/wm_Lao_2017.sav", user_na = TRUE)
wm <- as.data.frame(wm)
```

It is good to have a look at the data, even if it has  **`r ncol(wm)`** variables. 


## Select the variables that may be relevant 

First, we need to redefine some variable and adjust their right type (numeric, character, logic, etc.)

```{r}
# Rename and reformat some variables 
wm$age <- as.numeric(as.character(wm$WB4))

```


> We create a new data frame with only the variables for our analysis 

```{r}
#select the only relevant variables
df <- select(wm, 
             age, WAGEM,  hh6a, welevel, windex5, WM17, WAGEM, MA1, HH7, HH1, wmweight)
```

Then we need to keep only the observations corresponding to our definition: Women of age between 20 and 25.

```{r}
#Select only women ages between 20 and 24 and those completed the survey
df <-subset(df,  (WM17 == 1) & age >= 20 & age < 25)

```

That data frame has now **`r nrow(df)`** observations and  **`r ncol(df)`** variables 

# Descriptive statistics

```{r}
datasummary_skim(df, type = "numeric")
```

The variable `WAGEM` is the variable that will help us define  child marriage. But it has some missing information, so we should refine the data set. 

```{r}
df <- drop_na(df)
```

> The final data set has **`r nrow(df)`** observations (and  still `r ncol(df)` variables )

### Defening the variables with the right "type" 

```{r}
# Create a variable that records 1 for child marriage and 0 otherwise
df$Before18<- ifelse(df$WAGEM<18,1,0)

# That variable must be a factor variable
df$Before18 <- as.factor(df$Before18)

# the level of education is also a factor variable with 6 levels
df$welevel <- as.factor(df$welevel)

```

```{r}
datasummary_skim(df, type = "categorical")
```

## Cross-tabulations 
We can also do some cross-tabulations to asses the relations between child marriage and education  


```{r}
tab <- table(df$Before18, df$welevel)

# Column percentages
tt <- prop.table(tab, margin = ) * 100  # multiply by 100 for % 
kable(tt, digits =2, caption = "Child marriage vs Education")
```


```{r}
tab <- table(df$Before18, df$hh6a)

# Column percentages
tt <- prop.table(tab, margin = 2) * 100  # multiply by 100 for % 
kable(tt, digits =2, caption = "Child marriage vs Area")
```




## Logistic regression

We will use the DHS survey at the individual level and construct a binary response variable indicating whether there was marriage before 15 years old (or "*girl's marriage*"). We will try to understand which factor**S** may affect the probability to observe girl's marriage, and in particular whether, after controlling for the usual individual factors such as education, wealth and other socioeconomic factors, there is also an impact of environmental factors.   


```{r}
# Defining the variables of the model
Y<-"Before18"               # Response variable

XCovars <- c(1,4,5)   # age+education+wealth


formula_string<- paste(Y, paste(colnames(df)[XCovars], collapse=" + "), sep="~")
print(paste(" Regression formula: ",formula_string))

```

### Results {-}
We use a logit model with several explanatory variables

```{r, results='asis'}
# Logistics Regression
glm.fit <- glm(formula_string,
               data = df,
               family = binomial)

# Nice printing of the results (using paper and knitr packages)
pretty_lm2 <- prettify(summary(glm.fit))
kable(pretty_lm2, digits = 3)

```


### Confusion Matrix {-}

The confusion matrix shows how well the model predicted the outcome. If the model was perfect, we would only have elements on the diagonal of this matrix and 0’s everywhere. 

```{r, results=TRUE }
library("regclass")
cm <- confusion_matrix(glm.fit)
cm
```


```{r, results=TRUE }
TN <- cm[1,1]
FP <- cm[2,1]
FN <- cm[1,2]
TP <- cm[2,2]

# Some indicators 
misclass <- round(100* (FP+FN)/ (TN+TP+FN+FP), 1)

accuracy <- round(100* (TP + TN) /(TN+TP+FN+FP),1)
accuracy


```
In total, the model misclassified ( `r FP`   + `r FN` / `r nrow(df)`) = **`r misclass` %** of the outcome, which yields **an accuracy of `r accuracy` per cent** .

### Visual representation of the logistic model{-} 
We can also visualize the effect of some of the most significant variables in the model. 


```{r}
p_educ <- visreg(glm.fit, "welevel", scale="response", rug=0,  # for rugs =2
                 xlab="Education level",
                 ylab="Prob(Child Marriage)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()

p_educ

p_wealth <- visreg(glm.fit, "windex5", scale="response", rug=0,  # for rugs =2
                   xlab="Wealth index",
                   ylab="Prob(Child Marriage)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()

p_wealth

p_age <- visreg(glm.fit, "age", scale="response", rug=0,  # for rugs =2
                   xlab="Age (in 2017)",
                   ylab="Prob(Child Marriage)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()


# p_area <- visreg(glm.fit, "hh6a", scale="response", rug=0,  # for rugs =2
#                    xlab="Urban vs Rural",
#                    ylab="Prob(Child Marriage)", gg=TRUE) + 
#   ylim(0,1) +theme_minimal()
# 
# p_area

## Combining the graphics
figure <- ggarrange( p_educ, p_wealth, p_age, # p_area,
                     #labels = c("Edudation", "wealth"  "Age"),
                     ncol = 4, nrow = 1)

figure
```


We can see the marginal effects  of significant predictors in the model. 

As the level of education increases, the probability decreases. We can notice a clear drop in the probability between the category 3 "middle" and  4 “Fourth”. Also, all things being equal, the probability of child marriage decreases with the increase of the level of wealth. 

The probability of being married before 18 decreases also with age. This shows a lower prevalence of the practice in the past.

# But the analysis is not correct...
There is a sampling protocol and we should integrate the weights for each household. For that, we'll use the package  `survey`

```{r}
# Declaring survey sample design/weights and creating tables 
library(survey)
```

We have the find the variables that define the strata and the weights 

```{r}
wmdesign<-svydesign(id= ~HH1,strata= ~HH7, weights = ~wmweight, nest=TRUE, data=df)
```

 We can redo the cross-tabulation, taking into account the weights and stratas..
 
```{r }
library(RcmdrMisc)
# cross-tabulations (instead of correlation matrix)
colPercents(svytable(~Before18+welevel, wmdesign,round=TRUE))
```


```{r }
colPercents(svytable(~Before18+windex5, wmdesign,round=TRUE))
```

```{r}
# Full model with sampling design included

wm.sglm1 <- svyglm(formula = formula_string, 
                   design = wmdesign, 
                   family = binomial, 
                   data = df)

summary(wm.sglm1)
```


```{r}
# Visualizing regression

pnew_educ <- visreg(wm.sglm1, "welevel", scale="response", rug=0,  # for rugs =2
                 xlab="Education level",
                 ylab="Prob(Child Marriage)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()

pnew_educ

pnew_wealth <- visreg(wm.sglm1, "windex5", scale="response", rug=0,  # for rugs =2
                   xlab="Wealth index",
                   ylab="Prob(Child Marriage)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()

pnew_wealth

pnew_age <- visreg(wm.sglm1, "age", scale="response", rug=0,  # for rugs =2
                   xlab="Age (in 2017) ",
                   ylab="Prob(Child Marriage)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()

pnew_age


## Combining the graphics
New_figure <- ggarrange( pnew_educ, pnew_wealth, pnew_age,
                     #labels = c("Edudation", "wealth"  "Age"),
                     ncol = 3, nrow = 1)

New_figure


```



# Conclusion

The results suggest that: 

-  “*Education*”, 
- “*Age*”, 
and 
- “*wealth*” 

are important variables in explaining and predicting the outcome of marriage before 18 in Lao.

It is intuitive that “Education” and "wealth" empowers women and it significantly reduces the probability of being married before 18. 


Further research is needed to provide insights on these findings. 
