---
title: "Parallel Coordinates with R"
subtitle: "With Plotly"
author: 
  - Christophe Bontemps (SIAP)
output:
  html_document:
    df_print: paged
    toc: yes
    keep_md: yes
    code_folding: show
    fig_width: 6.5
    fig_height: 4
  pdf_document:
    df_print: kable
    toc: no
    keep_tex: yes
    fig_width: 6.5
    fig_height: 4
---

```{r Knitr_Global_Options, include=FALSE, cache=FALSE}
library(knitr)
opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE,
               autodep = TRUE, tidy = FALSE)
#opts_chunk$set(cache.rebuild=TRUE) 

# My colors:
SIAP.color <- "#0385a8"

```

`r if(knitr:::pandoc_to() == "latex") {paste("\\large")}` 


```{r packages, include=FALSE, echo = FALSE}
# Data management packages
library(jtools)
library(forcats)
library(tidyverse)

# Plotting packages
library(ggplot2)
library(RColorBrewer)
library(plotly)


# Nice presentation of results
library(papeR)
library(xtable)
library(kableExtra)

# Tables 
library(modelsummary)
library(gt) 
library(gtsummary)


```

# Introduction
This course is inspired by [this plotly web page](https://plotly.com/r/parallel-coordinates-plot/)

Creating parallel coordinates plots (PCP)  is **quite easy** if
- One use the right package (here *plotly*)
- The data are well structured

For example the famous *Iris* data set is well structured with observations in rows, and some variables as column. 

```{r Iris}
library(plotly)

IrisFlowers <- read.csv("https://raw.githubusercontent.com/bcdunbar/datasets/master/iris.csv")
```

Let's have a look on the data structure

```{r}
IrisFlowers %>% head()
```
 And in details, since all variables are numeric, we can see the number of observation,and think of *species_id* as a visual variable for colors. 
 
```{r}
datasummary_skim(IrisFlowers, type = "numeric")
```
 
### A simple PCP

```{r Data}
fig <- IrisFlowers %>% plot_ly(type = 'parcoords',
          line = list(color = ~species_id,
                      colorscale = list(c(0,'red'),c(0.5,'green'),c(1,'blue'))),
          dimensions = list(
            list(range = c(2,4.5),
                 label = 'Sepal Width', values = ~sepal_width),
            list(range = c(4,8),
                 constraintrange = c(5,6),
                 label = 'Sepal Length', values = ~sepal_length),
            list(range = c(0,2.5),
                 label = 'Petal Width', values = ~petal_width),
            list(range = c(1,7),
                 label = 'Petal Length', values = ~petal_length)
            )
          ) %>%
        layout(title = 'A simple Parallel Coordinates Plot')

fig

```


### More advanced for more complex data 
Let's have a look at some SDGs for many countries in the world

```{r}
library(plotly)
Mydata <- read.csv("Data/SDGperRegion.csv")
```


We have here `r nrow(Mydata)` observations (rows) and `r ncol(Mydata)` variables. 
 and 
```{r}
head(Mydata)
```



```{r}
Mydata <- Mydata %>%
  mutate(region = as.factor(region_code))


fig <- Mydata %>%
  plot_ly(width = 1000, height = 600) 
fig <- fig %>% add_trace(type = 'parcoords',
          line = list(color = ~region_code,
                       colorscale = "geyser", 
                     # colorscale = list(c(0,'red'),c(0.25,'pink'),
                    #                  c(0.5,'green'),c(0.75,'black'), c(1,'blue')), 
                     showscale = FALSE),
          dimensions = list(
            list(range = c(~min(population),~max(population)),
                 label = 'Population', values = ~population),
            list(range = c(~min(gross_national_income_atlas_method),
                           ~max(gross_national_income_atlas_method  )),
                 label = 'GNI', values = ~gross_national_income_atlas_method),
           list(range = c(~min(maternal_mortality_ratio   ),~max(maternal_mortality_ratio   )),
                 label = 'Maternal Mortality ratio', values = ~maternal_mortality_ratio   ), 
           list(range = c(~min( X.Tuberculosis..per.100k. ),
                           ~max( X.Tuberculosis..per.100k.   )),
                 label = 'Tuberculosis', values = ~ X.Tuberculosis..per.100k. ),
            list(range = c(~min(access_to_electricity),
                           ~max(access_to_electricity  )),
                 label = 'Electricity', values = ~access_to_electricity),
            list(range = c(~min(primary_completion_rate ),
                           ~max(primary_completion_rate   )),
                 label = 'Education ', values = ~primary_completion_rate ),
            list(range = c(~min(internet_use ),
                           ~max(internet_use   )),
                 label = 'internet_use ', values = ~internet_use )
           
            )
          ) %>%
        layout(title = 'A more complex Parallel Coordinates Plot')


fig

```

Obviously, there is a lot to optimize here, in particular the choice of the variables, the order of the axis, the colors, etc... But one can see how many dilensions can be visualized there. 







