---
title: "Exercise: How to Visualize 8 Numbers?"
author: "Christophe Bontemps (UN SIAP)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    toc: no
    keep_md: yes
    code_folding: hide
    code_download: true
    highlight: tango
    number_sections: yes
    theme: cerulean
  word_document: default
  pdf_document:
    keep_tex: yes
---


```{r setup, include=FALSE}
# Remember to disable png for Word export...
knitr::opts_chunk$set( message = FALSE, warning = FALSE, 
                       results = TRUE, echo = TRUE,
                       fig.width=7, fig.height=4, 
                       dev="png", 
                       dev.args=list(type="cairo"), dpi=96)

# My colors:
SIAP.color <- "#0385a8"

# Function used to recreate ggplot standard colors
ggplotColours <- function(n = 6, h = c(0, 360) + 15){
  if ((diff(h) %% 360) < 1) h[2] <- h[2] - 360/n
  hcl(h = (seq(h[1], h[2], length = n)), c = 100, l = 65)
}

```


```{r packages, include=FALSE, echo = FALSE}
# Data management packages
library(jtools)
library(forcats)
library(tidyverse)

# Plotting packages
library(ggplot2)
library(RColorBrewer)
library(plotly)


# Nice presentation of results
library(papeR)
library(xtable)
library(kableExtra)

# Tables 
library(modelsummary)
library(gt) 
library(gtsummary)
library(skimr)

```

Let’s consider an example using observational data on average teacher salaries (in 2010) and average total SAT scores for each of the 50 United States. The SAT (*Scholastic Aptitude Test*) is a high-stakes exam used for entry into college. 

> Are higher teacher salaries associated with better outcomes on the test at the state level? 

If so, should we adjust salaries to improve test performance? 

Let's look at the relation between SAT's scores (Y) and  Teachers Slaries (X). We can do a simple scatterplot of these data. We also fit a linear regression model.


```{r}
# From https://mdsr-book.github.io/mdsr3e/09-foundations.html#sec-confound
 load(file = 'data/SAT_2010.rda')


SAT_2010 <- SAT_2010 |>
  mutate(Salary = salary/1000)

SAT_plot <- ggplot(data = SAT_2010, aes(x = Salary, y = total)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  ylab("Average total score on the SAT") + 
  xlab("Average teacher salary (thousands of USD)") +
  theme_minimal()
SAT_plot
```
The regression associated  shows that the coeficient is clearly negative and significantly different from 0. 

```{r}
SAT_mod1 <- lm(total ~ Salary, data = SAT_2010)
broom::tidy(SAT_mod1)
```
```{r}
SAT_2010 |>
  skim(sat_pct)
```

The median is at  **`r median(SAT_2010$sat_pct)`**, so let's split the sample into two groups: high and low percentage of student who take the SAT. 

```{r}
# New variable created SAT_grp

SAT_2010 <- SAT_2010 |>
  mutate(SAT_grp = ifelse(sat_pct <= 27, "Low", "High"))

SAT_2010 |>
  group_by(SAT_grp) |>
  count()
```

```{r}
SAT_plot %+% SAT_2010 + 
  aes(color = SAT_grp) + 
  scale_color_brewer("% taking\nthe SAT", palette = "Set1")

```


```{r}
SAT_2010 |>
  group_by(SAT_grp) |>
  group_modify(~broom::tidy(lm(total ~ Salary, data = .x)))
```


```{r}
SAT_mod2 <- lm(total ~ Salary + sat_pct, data = SAT_2010)
broom::tidy(SAT_mod2)
```



```{r EXIT}
knitr::knit_exit()
```









```{r}
# Generating data with Simpson's paradox
set.seed(2512)

# Creating a dataset with three variables: X, Y, and Group
n <- 300
x <- rnorm(n, mean = 20, sd = 5)
y <- 3 * x + rnorm(n, mean = 0, sd = 10)
group <- factor(rep(1:3, each = n/3))

# Introducing Simpson's paradox
grouped_x <- split(x, group)
grouped_y <- split(y, group)

# Increasing the correlation within each subgroup
grouped_y[[1]] <- grouped_y[[1]] - 3 * grouped_x[[1]]* rnorm(n/3, mean = 1, sd = 3)
grouped_y[[2]] <- grouped_y[[2]] + 3 * grouped_x[[2]]* rnorm(n/3, mean = 1.2, sd = 3)
grouped_y[[3]] <- grouped_y[[3]] - 3 * grouped_x[[3]]* rnorm(n/3, mean = 1.5, sd = 3)

# Removing one data point from each group to have consistent number of rows
grouped_x <- lapply(grouped_x, function(x) x[-sample(length(x), 1)])
grouped_y <- lapply(grouped_y, function(y) y[-sample(length(y), 1)])

x <- unlist(grouped_x)
y <- unlist(grouped_y)

# Creating the final dataset
data <- data.frame(X = x, Y = y, Group = group[1:297])

```

# Stat des
```{r}
# Visualization of the paradox with regression lines and small multiples
# Plot with regression lines
ggplot(data, aes(x = X, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Simpson's Paradox",
       x = "X", y = "Y")
```


```{r}
# Visualization of the paradox with regression lines and small multiples
# Plot with regression lines
ggplot(data, aes(x = X, y = Y, color = Group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Simpson's Paradox",
       x = "X", y = "Y")
```

```{r}
# Plot with small multiples
ggplot(data, aes(x = X, y = Y, color = Group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Group) +
  labs(title = "Simpson's Paradox: Small Multiples",
       x = "X", y = "Y")

```


```{r}
# Generating data with Simpson's paradox in another way
set.seed(42)

# Creating a dataset with three variables: University, Applicants, and Admitted
university <- c(rep("University A", 100), rep("University B", 200))
applicants <- c(500, 200, 400, 100, 200, 150)
admitted <- c(250, 180, 350, 90, 100, 120)

# Creating the final dataset
data_another <- data.frame(University = university, Applicants = applicants, Admitted = admitted)

```


```{r}
# Calculate admission rates for each university
data_another$Admission_Rate <- data_another$Admitted / data_another$Applicants * 100
data_another

```

```{r}
# Adding gender information to the dataset
gender <- c(rep("Male", 200), rep("Female", 100))
data_another$Gender <- gender

# Calculate admission rates for each university and gender
admission_rates_by_gender <- aggregate(Admitted ~ University + Gender, data_another, sum)
applicants_by_gender <- aggregate(Applicants ~ University + Gender, data_another, sum)
data_another_gender <- merge(admission_rates_by_gender, applicants_by_gender, by = c("University", "Gender"))

data_another_gender$Admission_Rate_Gender <- data_another_gender$Admitted / data_another_gender$Applicants * 100
data_another_gender

```

## 2 

```{r}
# Generating data with Simpson's paradox in another way
set.seed(42)

# Creating a dataset with three variables: University, Applicants, and Admitted
university <- c(rep("University A", 100), rep("University B", 200))
applicants <- c(500, 200, 400, 100, 200, 150)
admitted <- c(250, 180, 350, 90, 100, 120)

# Creating the final dataset
data_another <- data.frame(University = university, Applicants = applicants, Admitted = admitted)

```



```{r}
# Visualization of the paradox with regression lines
ggplot(data_another, aes(x = Applicants, y = Admission_Rate, color = University)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Simpson's Paradox: Admission Rates by University",
       x = "Number of Applicants", y = "Admission Rate (%)") +
  theme_minimal()

```






####  New

```{r}
# Generating data with Simpson's paradox
set.seed(42)

# Creating a dataset with three variables: X, Y, and Group
n <- 100
x <- c(rnorm(n, mean = 10, sd = 2), rnorm(n, mean = 30, sd = 5))
y <- c(2 * x[1:n] + rnorm(n, mean = 0, sd = 2), 3 * x[(n+1):(2*n)] + rnorm(n, mean = 0, sd = 5))
group <- factor(rep(1:2, each = n))

# Creating the final dataset
data_simpson <- data.frame(X = x, Y = y, Group = group)

```

```{r}
# Visualization of the paradox with regression lines
ggplot(data_simpson, aes(x = X, y = Y, color = Group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Simpson's Paradox: X vs. Y",
       x = "X", y = "Y") +
  theme_minimal()

```

```{r}
# Visualization of the combined data with regression line
ggplot(data_simpson, aes(x = X, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Simpson's Paradox: Combined Data",
       x = "X", y = "Y") +
  theme_minimal()

```

```{r}
# Generating data with Simpson's paradox
set.seed(42)

# Creating a dataset with three variables: X, Y, and Group
n <- 100
x <- c(rnorm(n, mean = 10, sd = 2), rnorm(n, mean = 30, sd = 5))
y <- c(2 * x[1:n] + rnorm(n, mean = 0, sd = 2), 3 * x[(n+1):(2*n)] + rnorm(n, mean = 0, sd = 5))
group <- factor(rep(1:2, each = n))

# Creating the final dataset
data_simpson <- data.frame(X = x, Y = y, Group = group)

```


```{r}
# Visualization of the paradox with regression lines
ggplot(data_simpson, aes(x = X, y = Y, color = Group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Simpson's Paradox: X vs. Y",
       x = "X", y = "Y") +
  theme_minimal()

```

