---
title: "Live exercise with TAPOS 2024"
author: "Christophe"
date: "10-09-2024"
output:
  html_document:
    toc: yes
    code_folding: show
    df_print: paged
  word_document:
    toc: no
  pdf_document:
    toc: no
---


#  Data analysis of Trade data
We provide here some basic elements about the data set and descriptive statistics on some variables of interest.  
First we load some packages (libraries). 

> These libraries must have been installed beforefand using the bottom-right panel of RStudio. 

```{r}
library(tidyverse) 
```


* First we load the data
* Second we clean the data set imported from excel 
* Third we start analyzing the data and print a *summary*

```{r}
# Here I'm loading the data 
library(readxl)
MyData2 <- read_excel("Data/WV.1_Size_of_the_economy.xlsx")

```


We loaded the data set, in the importation from excel the first lines are quite problematic and we need some cleaning. 

## Cleaning 

```{r}
Temp <- MyData2[1:3, 9]

# Remove the first 3 lines 
Cleaned <- MyData2[4:308,]

# Extracting variables 3, 7and  5 
extract <- MyData2[,c(3,7,5)]
```

Now we have a real data set, we can view it on the top-right panel. 

 
### Change format of some variables

Most (if not all) variable are under the *character* format and this will be a problem for using some functions in the analysis. We would like to have *numerical* variables (format *numeric*).

We use the function *as.numeric* to change that and create a new variable (*NewPop*) to see if that affected the value and compare with the original variable  *Population* 

```{r}
# We transform the variable "population" to numeric
Cleaned <- Cleaned %>%
  mutate(NewPop = as.numeric(Population)) 
  
```

### Changing the ordering of variables 

We would like to have the two variables *Population* (character) and *NewPop* (Numeric) side by side at the beginning of the data frame *Cleaned*. We use the *select* function and the  *everything()* function to keep other variables where they were. 

```{r}
# Move specific columns to the front
Cleaned <- Cleaned %>%
  select(Population, NewPop, everything())
```

### Changing the name of some variables 
The variable with the name of the countries is currently "...1" which is not handy. Let's change that. 

```{r}
# Rename column "...1" to "Country"
Cleaned <- Cleaned %>%
  rename(Country = `...1`)
```

Reordering again 
 

```{r}
# Move specific columns to the front
Cleaned <- Cleaned %>%
  select(Country, everything())
``` 


> **Exercise**
Can you change the name of the variable ..8 to "PPP GNI per capita"? 



### Transforming all character variables into numeric using ChatGPT

Now, we used ChatGPT to find the write the code that will transform all character variables **but the first one** (which is the name of the countries ) into numeric.^[The prompt used is the following "*You are expert in R. Using tidyverse and the as.numeric function,  please transform all character variables into numeric for my data frame named Cleaned*". Then we add "*Good,  but keep the first one as character*"]

```{r}
# Transform all character variables to numeric, except the first column
Cleaned <- Cleaned %>%
  mutate(across(-1, ~ if(is.character(.)) as.numeric(.) else .))
  
```
The variable *NewPop* is not useful anymore since we changed the format of *Population* directly with that global command. Let us remove it: 

```{r}
Cleaned <- Cleaned %>%
  select(-NewPop)
```


## Summary statistics
Here are some summary statistics of one variable (here *Population*) 

```{r }
summary(Cleaned$Population)

```

### Using computed values in the text 
We can also compute several elements that we can use later, like the number of observations and the number of variables

```{r}
Nobs <- nrow(Cleaned)
Nvar <- ncol(Cleaned)
```

We can use that information directly in the text using mini-chunks and proudly (automatically) write that there are `r Nobs` observations and  `r Nvar` variables in the data set. 

# Analysis and graphics

##  A simple graphic

```{r}
library(ggplot2)

ggplot(Cleaned) +
  aes(y = Population) +
  geom_boxplot() +
  labs(title = "Horizontal Box Plot of Population",
       y = "Population") +
  coord_flip() +
  theme_minimal()
```


