---
title: "Working with Survey data"
author: "Christophe"
date: "29-08-2024"
output:
  html_document:
    toc: yes
    code_folding: show
    df_print: paged
  word_document:
    toc: no
  pdf_document:
    toc: no
---


#  Simple operations with Survey data
We provide here some basic elements about using survey data and computation of specific elements
First we load some packages (libraries). 

> These libraries must have been installed beforefand using the bottom-right panel of RStudio. 

```{r }
library(tidyverse) 
```

* First we load the data
* Second we compute some averages 
* Third we define new variables 

> For this exercise, we will generate synthetic data (simulations)

```{r warning=FALSE, include=FALSE}
# Set a seed for reproducibility
set.seed(123)
NbObs <- 200

# Create a data frame with 7 variables and 20 observations
household_survey_simulated <- tibble(
  household_id = 1:NbObs,                               
  household_size = sample(1:8, NbObs, replace = TRUE),  
  income = sample(20000:100000, NbObs, replace = TRUE),
  head_gender = sample(c("Male", "Female"), NbObs, replace = TRUE),  
  head_age = sample(18:80, NbObs, replace = TRUE),      
  location = sample(c("Urban", "Rural"), NbObs, replace = TRUE),  #
  random = rnorm(NbObs, 0, 1)               
) %>%
  mutate(weight = (random - min(random)) / 
                              (max(random) - min(random))) %>%
  select(-c(random))

write.csv2(household_survey_simulated, "Data/Household_Survey.csv")
```

## First we load the data
Here the data are on a server, online. 

```{r }
household_survey <- read.csv2("https://raw.githubusercontent.com/XtopheB/JICA23/main/Exercises/Household_Survey.csv")
```

Now we have a data set, we can view it on the top-right panel. 

Let's compute the average income for `rural` population . 

```{r}
# Step 1: Create a new data frame with only rural
household_rural <- household_survey %>%
  filter(location =="Rural" )

## NB This can be done with subset() too
household_rural_bis <- subset(household_survey, location =="Rural" )
```

So subsetting is filtering observations based on a logical test on some variables. 

```{r}
# Step 2: Calculate total income
total_rural  <- household_rural %>%
  summarize(total = sum(income))

# Step 3: Calculate the number of households 
Nb_rural <-  household_rural %>%
  count()

# Step 4: Compute the average income
avg_income <- total_rural$total / Nb_rural$n

# Display the result
print(avg_income)

```

So, our average income is `r avg_income`

## But Wait! We have to take into account the weights!!


```{r}
# Step 2: Calculate total weighted  income
total_weighted_rural  <- household_rural %>%
  mutate(weighted_income = income * weight) %>%
  summarize(total_weighted_income = sum(weighted_income))

# Step 3: Calculate the weighted number of households 
Nb_weighted_rural <-  household_rural %>%
  summarize(total_weights = sum(weight))
  
# Step 4: Compute the weighted average income
avg_weighted_income <- total_weighted_rural$total_weighted_income / Nb_weighted_rural$total_weights

print(avg_weighted_income)

```

## Or in a few line of code

```{r}
avg_weighted_income2 <- household_survey %>%
   filter(location =="Rural" ) %>%
  summarize(weighted_avg_income = sum(income * weight) / sum(weight)) %>%
   pull(weighted_avg_income) # This extract the only value of the data frame

print(avg_weighted_income2)
```
# Now we need to create a specific category 

> Task: Create a new column income_category that categorizes households into "Low",  or "High" income based on the value of income. Use the following thresholds:

* "Low" for income below $50,000
* "High" for income above $70,000

**Hint** Use the `ifelse`function 

```{r}
household_survey <- household_survey %>%
  mutate(income_category = ifelse(income < 50000, "Low", "High"))
```

## Can we do 3 categories? 
 
* "Low" for income below $30,000
* "Medium" for income between $30,000 and $70,000
* "High" for income above $70,000
 
```{r}
household_survey <- household_survey %>%
  mutate(income_category = ifelse(income < 30000, "Low",
                                   ifelse(income <= 70000, "Medium", "High")))
```
 
## Can we do advanced (complex)  cases?
> Task: Create a new column *special_status* that indicates whether a household should receive special consideration based on two conditions:

* The household_size is greater than 5.
* The income is below $30,000.

```{r}
household_survey <- household_survey %>%
  mutate(special_status = ifelse(household_size > 5 | income < 30000, 
                                  "Special Consideration", 
                                  "No Special Consideration"))
```

