---
title: "My title"
author: "Your name"
output:
  html_document:
    toc: no
    code_folding: show
    df_print: paged
  word_document:
    toc: no
  pdf_document:
    toc: no
---
# Your task: 

- Change the name (for your) in the YAML above
- Run the whole document using the "knit" button 
- Run the first chunk below, then the second
- Create a new chunk (No 1) where you'll count the number of observations (put that in Nobs)
- Create a new chunk (No 2) where you'll count the number of variable (put that in Nvar)
- Rerun the whole document
- Create a new chunk (No 3) where 
    - you will load new data using read.csv2("https://www.unsiap.or.jp/on_line/ML/MLData/ML_SouthAfrica.csv")
    - Create a scatterplot (ggplot) of ltexp vs FoodShr 
- Rerun the whole document


# Data analysis of Trade data
We provide here some basic elements about the data set and descriptive statistics on some variables of interest.  

* First we load the data
* second we print a summary

```{r}
TradeData <- read.csv("https://www.unsiap.or.jp/on_line/2024/RAP/Data/TradeData.csv")
```

We loaded the data set and compile the  report on `r Sys.Date()`

## Summary statistics
Here are some summary statistics of the export value (*Stat_Value* variable), all products combined:

```{r }
summary(TradeData$Stat_Value)
```

###  (new) chunk nb 1  will be here --- 


###   (new) chunk nb 2  will be here --- 

The data set contains   **`r  nrow(TradeData)`  observations**, and  **XXXX** variables.

Here is a list of all variables available in the data set: 

```{r}
names(TradeData)
```


The variable *Principle_Export* list all the types of goods exported from Vanuatu. There are **YYYY** different goods exported. Let's have the list of all these goods. 

```{r}
unique(TradeData$Principle_Exports)

```

# Now for something different: 

###  Create new chunk nb 3  here --- 


